---
title: "Step 5: differential Expression analysis for sub clusters"
date: "`r Sys.Date()`"
author: Yongjin Park
bibliography: "MS_Ref.bib"
output:
  html_document:
    toc: true
    keep_md: true
    self_contained: true
---


```{r include = F}
## unlink("result/step5",recursive = T)
## unlink("Fig/STEP5",recursive = T)
## unlink("Tab/DEG_mt*",recursive = T)
```

```{r GlobOpt, include=FALSE}
fig.dir = 'Fig/STEP5/'
knitr::opts_knit$set(eval.after = 'fig.cap')
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = fig.dir)
knitr::opts_chunk$set(fig.width=8, fig.height=8)
options(stringsAsFactors = FALSE)
FIG.CAP = '**Fig.**'
library(tidyverse)
library(data.table)
library(patchwork)
library(mmutilR)
library(kableExtra)
source("Util-rmd.R")
source("Util-alg.R")
.mkdir(fig.dir)
```

```{r read_hash}
.hash.hdr <- "result/step1/hash"
.hash.data <- fileset.list(.hash.hdr)
.hash.info <- read.hash(.hash.data)
.mkdir("result/step5/deg/")
```

```{r include = F}
sort.col <- function(.mat, .cts, .indvs){

    .fun <- function(ct) paste0(.indvs, "_", ct)

    .samples <-
        lapply(.cts, .fun) %>%
        do.call(what=c) %>%
        sort()

    .idx <- match(.samples, colnames(.mat))
    .idx <- .idx[!is.na(.idx)]
    .mat[, .idx, drop = FALSE]
}

combine.statistics <- function(.result){

    ret <- lapply(names(.result), function(v) {
        .mat <- .result[[v]]
        if(!is.matrix(.mat)) return(NULL)
        as.data.table(reshape2::melt(.mat, value.name = v))
    })

    out <- ret[[1]]
    for(j in 2:length(ret)){
        if(is.null(out)) {
            out <- ret[[j]]
        } else if(!is.null(ret[[j]])) {
            out <- merge(out, ret[[j]])
        }
    }
    return(out)
}

summarize.deg <- function(.dt, tot.cutoff = 3, disease.name = "MS", control.name = "HA") {

    ## adjust potential bias
    .dt[, b := 0]
    .dt[`tot` >= tot.cutoff, b := mean(cfa), by = .(gene, membership)]
    .dt[, cfa := cfa - b]

    .sd <- .dt[`tot` >= tot.cutoff,
               .(min.sd = sd(cfa) / sqrt(.N)),
               by = .(gene, membership)]

    .dt <- .dt %>%
        left_join(.sd) %>%
        filter(`tot` >= tot.cutoff) %>%
        mutate(w = 1/cfa.sd^2) %>%
        as.data.table

    ## average disease effect on the disease subjects
    ADD.dt <-
        .dt[disease == disease.name & `tot` >= tot.cutoff,
               .(ADD = sum(cfa * w) / sum(w),
                 ADD.se = sqrt(1/sum(w))),
               by = .(gene, membership)]

    ## average disease effect on the control subjects
    ADC.dt <-
        .dt[disease == control.name & `tot` >= tot.cutoff,
               .(ADC = -sum(cfa * w)/ sum(w),
                 ADC.se = sqrt(1/sum(w))),
               by = .(gene, membership)]

    ## average disease effect
    ADE.dt <-
        .dt[`tot` >= tot.cutoff] %>%
        mutate(cfa.s = if_else(disease == disease.name, 1, -1) * cfa) %>%
        mutate(tot.s = if_else(disease == disease.name, 1, -1) * tot) %>%
        as.data.table %>%
        (function(xx) {
            xx[,
               .(ADE = sum(cfa.s * w) / sum(w),
                 ADE.se = sqrt(1/sum(w)),
                 tot = sum(tot)),
               by = .(gene, membership)
               ]
        }) %>%
        mutate(z = ADE/ADE.se) %>%
        mutate(pv = 2 * pnorm(abs(z), lower.tail=FALSE)) %>%
        group_by(membership) %>% 
        mutate(fwer = p.adjust(pv, "holm")) %>%
        mutate(fdr = p.adjust(pv, "fdr")) %>%
        ungroup() %>% 
        as.data.table

    left_join(ADE.dt, ADD.dt) %>% left_join(ADC.dt) %>% as.data.table
}

format.deg.tab <- function(deg.tab, fig.hdr){

    .deg.cast <- function(.dt, .var, .fun = mean){
        .dt %>%
            mutate(k = .var %&% "_" %&% membership) %>% 
            dcast(ensembl_gene_id + hgnc_symbol ~ k,
                  value.var = .var,
                  fun.aggregate=.fun)
    }

    ret <-
        .deg.cast(deg.tab, "z", .fun = function(x) num.round(mean(x,na.rm=T))) %>%
        left_join(.deg.cast(deg.tab, "pv", .fun = function(x) num.sci(min(x,na.rm=T)))) %>% 
        left_join(.deg.cast(deg.tab, "fdr", .fun = function(x) num.sci(min(x,na.rm=T)))) %>% 
        as.data.table()

    ret[, .file := fig.hdr %&%
              ensembl_gene_id %&% "_" %&%
              hgnc_symbol %&% ".pdf"]

    ret[, .link := "[PDF](" %&% .file %&% ") "]
    ret[, .file := NULL]

    ret[!str_starts(ensembl_gene_id, "ENSG"), ensembl_gene_id := ""]

    return(ret)
}
```

```{r major_level_deg}
major.deg.tab <- fread("Tab/DEG_MS_vs_HC.txt.gz")
```

```{r read_bulk_deg_results}
## read.bulk <- function(.file) {
##     fread(.file) %>%
##         rename(hgnc_symbol = gene_name) %>%
##         rename(bulk.t = t, bulk.pv = P.Value, bulk.qv = adj.P.Val) %>%
##         dplyr::select(hgnc_symbol, starts_with("bulk"))
## }
## .file <- "data/DEG/20180513/deg.ms.hc.treg.mem.exvivo.sex_covar.ruv.txt"
## bulk.mTreg.dt <- read.bulk(.file)
## .file <- "data/DEG/20180513/deg.ms.hc.teff.mem.exvivo.sex_covar.ruv.txt"
## bulk.mTconv.dt <- read.bulk(.file)
```

# 1. mTconv subtype DEG analysis

```{r mtconv_read_annot}
annot.dt <-
    fread("Tab/step4_mtconv_leiden.txt.gz") %>%
    left_join(.hash.info) %>%
    na.omit()
```

```{r}
nnz.cutoff <- 500
cv.cutoff <- 1.0
```

```{r mtconv_estimate_deg_stat}
.data <- fileset.list("result/step1/matrix")
.deg.data <- fileset.list("result/step5/deg/mtconv_hc_ms")

if.needed(.deg.data, {
    .deg.data <-
        rcpp_mmutil_copy_selected_columns(.data$mtx,
                                          .data$row,
                                          .data$col,
                                          unique(annot.dt$tag),
                                          "result/step5/deg/mtconv_hc_ms")
})

.scores <- rcpp_mmutil_compute_scores(.deg.data$mtx,
                                      .deg.data$row,
                                      .deg.data$col)

row.scores <- setDT(.scores$row) %>%
    rename(gene = name) %>%                      # 
    filter(!str_detect(`gene`,"[Hh]ashtag")) %>% # remove hashtag
    as.data.table() %>%
    parse.gene()

qc.features <- row.scores[nnz > nnz.cutoff & cv > cv.cutoff]

.qc.hdr <- "result/step5/deg/qc_mtconv_hc_ms"
.qc.data <- fileset.list(.qc.hdr)

if.needed(.qc.data, {
    .qc.data <-
        rcpp_mmutil_copy_selected_rows(.deg.data$mtx,
                                       .deg.data$row,
                                       .deg.data$col,
                                       qc.features$gene,
                                       .qc.hdr)
})

.file <- "result/step5/deg/mtconv_hc_ms.rds"

if.needed(.file, {

    .membership <-
        annot.dt[, .(tag, membership)] %>%
        as.data.frame()

    .cell2indv <- annot.dt[, .(tag, subject)] %>%
        unique %>%
        as.data.frame()

    .indv2exp <- .cell2indv %>%
        select(subject) %>%
        mutate(disease = substr(`subject`, 1, 2)) %>%
        as.data.frame()

    .deg.stat <-
        make.cocoa(.qc.data, .membership, .cell2indv, .indv2exp,
                   knn = 100, .rank = 15, .take.ln = TRUE,
                   impute.by.knn = TRUE, num.threads = 16)

   saveRDS(.deg.stat, .file)
})
.deg.stat <- readRDS(.file)

.cts <- unique(annot.dt$membership)
.indvs <- unique(annot.dt$subject)

.hc.ms.dt <-
    list(tot = sort.col(.deg.stat$sum, .cts, .indvs),
         cfa = sort.col(.deg.stat$resid.ln.mu, .cts, .indvs),
         cfa.sd = sort.col(.deg.stat$resid.ln.mu.sd, .cts, .indvs)) %>%
    combine.statistics() %>%
    na.omit() %>%
    as.data.table() %>% 
    (function(x) {
        x[, c("sample", "membership") := tstrsplit(as.character(Var2), split="_")];
        x[, disease := substr(`sample`, 1, 2)];
        x[, gene := as.character(Var1)];
        x
    }) %>% 
    dplyr::select(-Var1, -Var2) %>% 
    as.data.table()

hc.ms.deg <-
    summarize.deg(.hc.ms.dt, tot.cutoff = 5) %>%
    parse.gene() %>%
    as.data.table()
```

```{r mtconv_ms_vs_hc_results, results="asis", echo = FALSE}
.mkdir("Tab")
.file <- "Tab/DEG_mtconv_MS_vs_HC.txt.gz"

if.needed(.file, {
    fwrite(hc.ms.deg, .file, sep="\t")
})
cat("[**DOWNLOAD:** mTconv DEG MS vs HC]("%&% .file %&% ")\n\n")
```

### Found `r num.int(nrow(unique(hc.ms.deg[fdr < .1, .(gene)])))` unique genes strongly perturbed by MS with FDR 10%

* Up-regulated: `r num.int(nrow(unique(hc.ms.deg[fdr < .1 & z > 0, .(gene)])))`

* Down-regulated:  `r num.int(nrow(unique(hc.ms.deg[fdr < .1 & z < 0, .(gene)])))`

* Total pairs of genes and clusters: `r num.int(nrow(hc.ms.deg))`

```{r}
count.deg <- function(.dt, fdr.cutoff = .1) {
    .dt[fdr < fdr.cutoff &
        sign(ADD) == sign(ADE) &
        sign(ADC) == sign(ADE),
        .(n = .N),
        by = .(membership,
               direction = if_else(z > 0, "up", "down"))
        ] %>%
        mutate(direction = factor(direction, c("up", "down"))) %>%
        group_by(membership) %>%
        arrange(desc(direction)) %>%
        mutate(nc = cumsum(n)) %>%
        ungroup
}
```

```{r Fig_mTconv_DEG_count, fig.width=2, fig.height=2, echo = FALSE, results="asis"}
.df <- count.deg(hc.ms.deg)

plt <-
    .gg.plot(.df, aes(x=factor(membership, 1:50), y=n, group=direction, fill=direction)) +
    theme(legend.position = c(1,1)) +
    theme(legend.justification = c(1,1)) +
    xlab("cluster") + ylab("# DEG") +
    geom_bar(stat="identity") +
    geom_text(aes(y=nc, label=n)) +
    scale_fill_manual(values = c("#ef8a62","#67a9cf"))

print(plt)
.file <- fig.dir %&% "/Fig_mTconv_DEG_count.pdf"
.gg.save(filename = .file, plot = plt, width=2, height=2)
```

#### Examples

```{r include = F}
show.each.gene <- function(g, .data, .stat, .cutoff = 3) {

    .dt.g <- na.omit(.data[gene == g]) %>% filter(tot > .cutoff)
    .stat.dt <- na.omit(.stat[gene == g]) %>% rename(total = tot)

    .dir <- .stat.dt[which.max(abs(.stat.dt$ADE)), .(ADE)] %>%
        unlist

    if(.dir > 0) {

        .dt <- .dt.g %>%
            left_join(.stat.dt, by = c("membership", "gene")) %>%
            group_by(disease, membership) %>%
            arrange(cfa) %>%
            mutate(j = 1:n()) %>%
            ungroup() %>%
            mutate(x = disease %&% "_" %&% j)

    } else {

        .dt <- .dt.g %>%
            left_join(.stat.dt, by = c("membership", "gene")) %>%
            group_by(disease, membership) %>%
            arrange(desc(cfa)) %>%
            mutate(j = 1:n()) %>%
            ungroup() %>%
            mutate(x = disease %&% "_" %&% j)

    }

    .dt <- .dt %>%
        mutate(.pv = if_else(is.na(pv), "1", num.sci(pv))) %>%
        mutate(membership = str_pad(as.integer(membership),2,pad=0)) %>% 
        mutate(.facet = membership %&% "\n(" %&% .pv %&% ")") %>%
        mutate(ADD.z = ADD/ADD.se, ADC.z = ADC/ADC.se) %>%
        as.data.table

    .stat.dt <- .dt[, .(membership,
                        ADE, ADE.se,
                        ADD, ADD.se,
                        ADC, ADC.se,
                        pv, .facet)] %>%
        unique

    .aes <- aes(x = x,
                y = cfa,
                ymin = cfa - cfa.sd,
                ymax = cfa + cfa.sd,
                fill = disease)

    .scale.y <- scale_y_continuous(labels = function(x) num.round(exp(x)))

    p1 <-
        .gg.plot(.dt, .aes) +
        theme(legend.position = "top") +
        facet_wrap(~.facet, nrow = 1, scales="free") +
        .scale.y +
        theme(axis.text.x = element_text(size=4)) +
        theme(axis.title.x = element_blank()) +
        geom_hline(yintercept = 0, linewidth = 1, colour = "gray") +
        geom_hline(data = .stat.dt, aes(yintercept = ADE), linewidth = 1, colour = "#F8766D") +
        geom_linerange(linewidth = .2) +
        geom_point(aes(size = tot), pch = 21, stroke = .2) +
        scale_size_continuous("total(k)", range=c(0.1, 2.5), labels = function(x) num.int(x/1e3)) +
        scale_fill_manual(values = c("gray90","#F8766D")) +
        xlab("") + ylab("fold change") + ggtitle(g)

    .temp.1 <- .stat.dt %>%
        dplyr::select(membership, pv, ADE, ADD, ADC) %>%
        mutate(ADC = - ADC) %>%                  # flip the effect
        melt(id.vars=c("membership", "pv"), value.name = "mu")

    .temp.2 <- .stat.dt %>%
        dplyr::select(membership, ADE.se, ADD.se, ADC.se) %>%
        melt(id.vars="membership", value.name = "se") %>%
        mutate(variable = str_remove(variable, ".se"))

    .temp <- left_join(.temp.1, .temp.2, by = c("membership", "variable"))

    .aes <- aes(x = variable,
                y = mu,
                ymin = mu - 2 * se,
                ymax = mu + 2 * se)

    p2 <-
        .gg.plot(.temp, .aes) +
        theme(axis.text.x = element_text(angle=90, hjust = 1, vjust= .5)) +
        facet_wrap(~membership, nrow = 1, scales="free") +
        .scale.y + ylab("fold change") + xlab("") +
        geom_hline(yintercept = 0, lty = 2, linewidth = .5, colour = "black") +
        geom_linerange(linewidth = .2) +
        geom_point(aes(fill = variable, stroke = .2, size = -log10(pv)), pch = 21) +
        scale_fill_brewer(palette = "Greens", guide="none") +
        scale_size_continuous("p-value", range=c(.5, 2.5), breaks=seq(0, 10, 2), labels = function(x) 10^(-x))

    plt <- (p1 / p2) + patchwork::plot_layout(heights = c(4,2))
    return(plt)
}
```

```{r}
hc.ms.deg[, c("ensembl_gene_id", "hgnc_symbol") := tstrsplit(`gene`, split="_")]

major.deg <- major.deg.tab[fwer < .05]$hgnc_symbol

.genes.show <-
    hc.ms.deg[hgnc_symbol %in% major.deg &
              sign(ADD) == sign(ADE) &
              sign(ADC) == sign(ADE)] %>%
    select(gene) %>%
    unique() %>%
    unlist() %>%
    as.character()
```

```{r Fig_mTconv_DEG_example, results="asis", echo = FALSE}
.mkdir(fig.dir %&% "/example/")
for(g in .genes.show){
    .file <- fig.dir %&% "/example/Fig_mTconv_DEG_example_" %&% g %&% ".pdf"
   if.needed(.file, {
       plt <- show.each.gene(g, .hc.ms.dt, hc.ms.deg)
       .gg.save(filename = .file, cat.link = FALSE, plot = plt, width=4, height=3)
   })
}
```

```{r Tab_mTconv_DEG_FDR, results="asis", echo = FALSE}
fig.hdr <- fig.dir %&% "/example/Fig_mTconv_DEG_example_"
deg.tab <- hc.ms.deg[gene %in% .genes.show]

format.deg.tab(deg.tab, fig.hdr) %>% 
    kbl() %>% 
    kable_paper("striped", full_width=F, html_font = "Helvetica") %>%
    column_spec(1, bold = T)
cat("\n\n\n")
```


# 2. mTreg subtype DEG analysis

```{r mtreg_read_annot}
annot.dt <-
    fread("Tab/step4_mtreg_leiden.txt.gz") %>%
    left_join(.hash.info) %>%
    na.omit()
```

```{r mtreg_estimate_deg_stat}
.mkdir("result/step5/deg/")
.data <- fileset.list("result/step1/matrix")
.deg.data <- fileset.list("result/step5/deg/mtreg_hc_ms")
.mkdir(dirname(.deg.data$mtx))

if.needed(.deg.data, {
    .deg.data <-
        rcpp_mmutil_copy_selected_columns(.data$mtx,
                                          .data$row,
                                          .data$col,
                                          unique(annot.dt$tag),
                                          "result/step5/deg/mtreg_hc_ms")
})

.scores <- rcpp_mmutil_compute_scores(.deg.data$mtx,
                                      .deg.data$row,
                                      .deg.data$col)

row.scores <- setDT(.scores$row) %>%
    rename(gene = name) %>%                      # 
    filter(!str_detect(`gene`,"[Hh]ashtag")) %>% # remove hashtag
    as.data.table() %>%
    parse.gene()

qc.features <- row.scores[nnz > nnz.cutoff & cv > cv.cutoff]

.qc.hdr <- "result/step5/deg/qc_mtreg_hc_ms"
.qc.data <- fileset.list(.qc.hdr)

if.needed(.qc.data, {
    .qc.data <-
        rcpp_mmutil_copy_selected_rows(.deg.data$mtx,
                                       .deg.data$row,
                                       .deg.data$col,
                                       qc.features$gene,
                                       .qc.hdr)
})

.file <- "result/step5/deg/mtreg_hc_ms.rds"

if.needed(.file, {

    .membership <-
        annot.dt[, .(tag, membership)] %>%
        as.data.frame()

    .cell2indv <- annot.dt[, .(tag, subject)] %>%
        unique %>%
        as.data.frame()

    .indv2exp <- .cell2indv %>%
        select(subject) %>%
        mutate(disease = substr(`subject`, 1, 2)) %>%
        as.data.frame()

    .deg.stat <-
        make.cocoa(.qc.data, .membership, .cell2indv, .indv2exp,
                   knn = 100, .rank = 15, .take.ln = TRUE,
                   impute.by.knn = TRUE, num.threads = 16)

   saveRDS(.deg.stat, .file)
})
.deg.stat <- readRDS(.file)

.cts <- unique(annot.dt$membership)
.indvs <- unique(annot.dt$subject)

.hc.ms.dt <-
    list(tot = sort.col(.deg.stat$sum, .cts, .indvs),
         cfa = sort.col(.deg.stat$resid.ln.mu, .cts, .indvs),
         cfa.sd = sort.col(.deg.stat$resid.ln.mu.sd, .cts, .indvs)) %>%
    combine.statistics() %>%
    na.omit() %>%
    as.data.table() %>% 
    (function(x) {
        x[, c("sample", "membership") := tstrsplit(as.character(Var2), split="_")];
        x[, disease := substr(`sample`, 1, 2)];
        x[, gene := as.character(Var1)];
        x
    }) %>% 
    dplyr::select(-Var1, -Var2) %>% 
    as.data.table()

hc.ms.deg <-
    summarize.deg(.hc.ms.dt, tot.cutoff = 5) %>%
    parse.gene()
```

```{r mtreg_ms_vs_hc_results, results="asis", echo = FALSE}
.mkdir("Tab")
.file <- "Tab/DEG_mtreg_MS_vs_HC.txt.gz"

if.needed(.file, {
    fwrite(hc.ms.deg, .file, sep="\t")
})
cat("[**DOWNLOAD:** mtreg DEG MS vs HC]("%&% .file %&% ")\n\n")
```

### Found `r num.int(nrow(unique(hc.ms.deg[fdr < .1, .(gene)])))` unique genes strongly perturbed by MS with FDR 10%

* Up-regulated: `r num.int(nrow(unique(hc.ms.deg[fdr < .1 & z > 0, .(gene)])))`

* Down-regulated:  `r num.int(nrow(unique(hc.ms.deg[fdr < .1 & z < 0, .(gene)])))`

* Total pairs of genes and clusters: `r num.int(nrow(hc.ms.deg))`

```{r Fig_mTreg_DEG_count, fig.width=2, fig.height=2, echo = FALSE, results="asis"}
.df <- count.deg(hc.ms.deg)

plt <-
    .gg.plot(.df, aes(x=factor(membership, 1:50), y=n, group=direction, fill=direction)) +
    theme(legend.position = c(1,1)) +
    theme(legend.justification = c(1,1)) +
    xlab("cluster") + ylab("# DEG") +
    geom_bar(stat="identity") +
    geom_text(aes(y=nc, label=n)) +
    scale_fill_manual(values = c("#ef8a62","#67a9cf"))

print(plt)
.file <- fig.dir %&% "/Fig_mtreg_DEG_count.pdf"
.gg.save(filename = .file, plot = plt, width=2, height=2)
```

```{r}
hc.ms.deg[, c("ensembl_gene_id", "hgnc_symbol") := tstrsplit(`gene`, split="_")]

major.deg <- major.deg.tab[fwer < .05]$hgnc_symbol

.genes.show <-
    hc.ms.deg[hgnc_symbol %in% major.deg &
              sign(ADD) == sign(ADE) &
              sign(ADC) == sign(ADE)] %>%
    select(gene) %>%
    unique() %>%
    unlist() %>%
    as.character()
```

```{r Fig_mTreg_DEG_example, results="asis", echo = FALSE}
.mkdir(fig.dir %&% "/example/")
for(g in .genes.show){
    .file <- fig.dir %&% "/example/Fig_mTreg_DEG_example_" %&% g %&% ".pdf"
    if.needed(.file, {
        plt <- show.each.gene(g, .hc.ms.dt, hc.ms.deg)
        .gg.save(filename=.file, cat.link=FALSE, plot=plt, width=4, height=3)
    })
}
cat("\n\n\n")
```

```{r Tab_mTreg_DEG_FDR, results="asis", echo = FALSE}
fig.hdr <- fig.dir %&% "/example/Fig_mTreg_DEG_example_"
deg.tab <- hc.ms.deg[gene %in% .genes.show]

format.deg.tab(deg.tab, fig.hdr) %>% 
    kbl() %>% 
    kable_paper("striped", full_width=F, html_font = "Helvetica") %>%
    column_spec(1, bold = T)
cat("\n\n\n")
```

