---
title: "Merge data sets and perform simple Q/C"
author: Yongjin Park
theme: jekyll-theme-minimal
date: "`r Sys.Date()`"
bibliography: "reference.bib"
---

```{r GlobOpt, include = FALSE}
################################################################
fig.dir = 'Fig/STEP1/'
dir.create(fig.dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_knit$set(eval.after = 'fig.cap')
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = fig.dir)
knitr::opts_chunk$set(fig.width=8, fig.height=8)
options(stringsAsFactors = FALSE)
FIG.CAP = '**Fig.**'

library(tidyverse)
library(data.table)
library(patchwork)
source("Util-rmd.R")
.fread <- function(...) fread(..., header=FALSE)
.fwrite <- function(...) fwrite(..., quote=FALSE, sep="\t", row.names = FALSE, col.names = FALSE)
.read.mat <- function(...) .fread(...) %>% as.matrix
.read.vec <- function(...) .fread(...) %>% unlist(use.names=FALSE)
```


# Outline

1. Merge five batches into one file set

2. Remove cells with too few counts and a high mitochondrial fraction

# 1. Merge all the available batches into one file set

## Combine both protein and RNA features

```{bash merging_data, message=FALSE}
mkdir -p result/step1/

[ -f result/step1/features.tsv.gz ] || \
	cp data/features.tsv.gz result/step1/features.tsv.gz

[ -f result/step1/matrix.mtx.gz ] || \
	mmutil_merge_col \
		result/step1/features.tsv.gz 0 result/step1/matrix \
		data/batch_3.mtx.gz \
		data/batch_3.feature.tsv.gz \
		data/batch_3.barcode.tsv.gz \
		data/batch_4.mtx.gz \
		data/batch_4.feature.tsv.gz \
		data/batch_4.barcode.tsv.gz \
		data/batch_5.mtx.gz \
		data/batch_5.feature.tsv.gz \
		data/batch_5.barcode.tsv.gz \
		data/batch_6.mtx.gz \
		data/batch_6.feature.tsv.gz \
		data/batch_6.barcode.tsv.gz \
		data/batch_7.mtx.gz \
		data/batch_7.feature.tsv.gz \
		data/batch_7.barcode.tsv.gz \
		2> /dev/null

[ -f result/step1/matrix.cols.gz ] || \
    gzip -cd result/step1/matrix.columns.gz | \
		awk '{ print $1 "_" ($2 + 2) }' | \
		gzip -c > result/step1/matrix.cols.gz

[ -f result/step1/matrix.batch.gz ] || \
	gzip -cd result/step1/matrix.cols.gz | \
		awk -F '_' '{ print $2 }' | \
		gzip -c > result/step1/matrix.batch.gz
```

## Take RNA-seq data with basic Q/C steps

```{sh take_rna_data, message=FALSE}
if ! [ -f result/step1/rna.mtx.gz ] ; then

	gzip -cd result/step1/matrix.rows.gz | \
		awk '!/anti_/ && !/MT-/' | \
		gzip > result/step1/rna.genes.gz

	mmutil_select_row \
		result/step1/matrix.mtx.gz \
		result/step1/matrix.rows.gz \
		result/step1/matrix.cols.gz \
		result/step1/rna.genes.gz \
		result/step1/rna 2> /dev/null

	mmutil_index_col result/step1/rna.mtx.gz 2> /dev/null
fi
```



```{sh take_mt_genes, message=FALSE}
if ! [ -f result/step1/mt.mtx.gz ] ; then

	gzip -cd result/step1/matrix.rows.gz | \
		awk '/MT-/' | \
		gzip -c > result/step1/mt.genes.gz

	mmutil_select_row \
		result/step1/matrix.mtx.gz \
		result/step1/matrix.rows.gz \
		result/step1/matrix.cols.gz \
		result/step1/mt.genes.gz \
		result/step1/mt 2> /dev/null

	mmutil_index_col result/step1/mt.mtx.gz 2> /dev/null

fi
```


```{sh compute_cell_score, message=FALSE}
[ -f result/step1/rna.cell_score.gz ] || \
	mmutil_col_score \
		result/step1/rna.mtx.gz \
		result/step1/rna.cols.gz \
		result/step1/rna.cell_score.gz 2> /dev/null

[ -f result/step1/mt.cell_score.gz ] || \
	mmutil_col_score \
		result/step1/mt.mtx.gz \
		result/step1/mt.cols.gz \
		result/step1/mt.cell_score.gz 2> /dev/null
```


```{sh filter_cells_bbknn_svd}
[ -d result/step1/bbknn ] || mkdir -p result/step1/bbknn

[ -f result/step1/rna.batch.gz ] || \
	gzip -cd result/step1/rna.cols.gz | \
		awk -F '_' '{ print $2 }' | \
		gzip -c > result/step1/rna.batch.gz

[ -f result/step1/bbknn/rna.mtx.gz ] || \
	mmutil_bbknn --mtx result/step1/rna.mtx.gz \
				 --col result/step1/rna.cols.gz \
				 --batch result/step1/rna.batch.gz \
				 --knn 100 --rank 3 --log_scale \
				 --out result/step1/bbknn/rna 2> /dev/null
```

# 2. Cell Q/C by the number of non-zero elements and mictochondrial activities

```{r cell_score_cutoff}
.read.score <- function(...) {
    ret = .fread(...,
                 col.names=c("cell", "nnz", "ntot", ".mean", ".sd", ".cv"))

    return(ret)
}

.dt.rna <- .read.score("result/step1/rna.cell_score.gz")
.dt.mt <- .read.score("result/step1/mt.cell_score.gz")

.file <- "result/step1/qc_table.txt.gz"

if(!file.exists(.file)) {

    .qc.dt <- left_join(.dt.rna, .dt.mt, by="cell", suffix=c(".nonmt", ".mt")) %>%
        mutate(tot.mt = ntot.mt * .mean.mt) %>%
        mutate(tot.nonmt = ntot.nonmt * .mean.nonmt) %>%
        mutate(tot = tot.nonmt + tot.mt) %>%
        mutate(mito.frac = 100 * tot.mt / tot)

    .factors <- .read.mat("result/step1/bbknn/rna.factors.gz") %>% scale

    .temp <- .factors %>%
        as.data.table %>%
        (function(x) { colnames(x) <- "PC" %&% 1:ncol(x); x })

    .qc.dt <- cbind(.qc.dt, .temp) %>% as.data.table %>% na.omit

    set.seed(1)

    .x <- .qc.dt$mito.frac
    .kmeans <- kmeans(.x, centers=2, nstart=100)
    k.valid <- which.max(.kmeans$size)
    mito.cutoff <- max(.x[.kmeans$cluster == k.valid])

    .y <- .qc.dt$nnz.nonmt
    .kmeans <- kmeans(log10(.y), centers=2, nstart=100)
    k.valid <- which.max(.kmeans$size)
    nnz.cutoff <- max(.y[.kmeans$cluster != k.valid])

    .dat <- .qc.dt[, .(PC1, PC2, PC3)] %>% as.matrix
    pc.valid <- apply(scale(.dat), 1, function(x) all(abs(x) < 3)) %>%
        which

    .qc.dt <- .qc.dt %>%
        mutate(j = 1:n()) %>%
        mutate(qc = if_else(nnz.nonmt > nnz.cutoff &
                            mito.frac < mito.cutoff &
                            j %in% pc.valid,
                            "pass",
                            "fail")) %>%
        dplyr::select(-j) %>%
        as.data.table

    fwrite(.qc.dt, .file, col.names = TRUE, row.names = FALSE)
} else {
    .qc.dt <- fread(.file, header = TRUE)
}
```

```{r qc_colours}
.qc.cols <- c("#8856a7", "gray70")
```

```{r fig_cell_nnz, fig.width=6, fig.height=3}
plt <-
    .gg.plot(.qc.dt, aes(x=nnz.nonmt)) +
    geom_histogram(aes(fill=qc), bins=100, colour="black", size=.1) +
    scale_x_continuous("number of genes detected in each cell") +
    scale_fill_manual(values = .qc.cols) +
    ylab("number of cells") +
    theme(legend.position="none")

print(plt)
.file <- fig.dir %&% "/Fig_cell_nnz.pdf"
.gg.save(filename = .file, plot = plt, width=6, height=3)
```

```{r results="asis", echo = FALSE}
cat("[PDF](" %&% .file %&% ")")
```

```{r fig_cell_mito, fig.width=6, fig.height=3}
.lab <- function(x) num.round(pmax(0,exp(x)-1))

plt <-
    .gg.plot(.qc.dt, aes(x=log(1 + mito.frac))) +
    geom_histogram(aes(fill=qc), bins=100, colour="black", size=.1) +
    scale_fill_manual(values = .qc.cols) +
    scale_x_continuous("% mitochondrial activity", labels = .lab) +
    ylab("count of cells") +
    theme(legend.position="none")

print(plt)
.file <- fig.dir %&% "/Fig_cell_mito.pdf"
.gg.save(filename = .file, plot = plt, width=6, height=3)
```

```{r results="asis", echo = FALSE}
cat("[PDF](" %&% .file %&% ")")
```

* Keep `r num.int(nrow(.qc.dt[qc == "pass"]))` cells, discard `r num.int(nrow(.qc.dt[qc == "fail"]))` cells

```{r Fig_svd_batch, fig.width=2, fig.height=2, echo = FALSE}
.temp <- .qc.dt[qc == "pass"]
.temp[, c("barcode","batch") := tstrsplit(cell, split="_")]

plt <-
    .gg.plot(.temp[sample(nrow(.temp))], aes(x=PC1, y=PC2)) +
    geom_point(aes(fill=batch), size = 1, pch = 21, stroke=0, show.legend=FALSE) +
    scale_fill_brewer(palette = "Spectral")

print(plt)
.file <- fig.dir %&% "/Fig_svd_batch.pdf"
.gg.save(filename = .file, plot = plt, width=2, height=2)
```

```{r results="asis", echo = FALSE}
cat("[PDF](" %&% .file %&% ")")
```

## Matrix data after Q/C: `r num.int(nrow(.qc.dt[qc == "pass"]))` cells

```{sh select_qc_cells}
[ -f result/step1/matrix_qc.select.gz ] || \
	gzip -cd result/step1/qc_table.txt.gz | \
		awk -F',' '$NF == "pass" { print $1 }' | \
		gzip -c > result/step1/matrix_qc.select.gz

[ -f result/step1/matrix_qc.mtx.gz ] || \
	mmutil_select_col \
		result/step1/matrix.mtx.gz \
		result/step1/matrix.cols.gz \
		result/step1/matrix_qc.select.gz \
		result/step1/matrix_qc 2> /dev/null
```

## Gene Q/C to remove genes with too many zeros

```{sh gene_qc_rna_data}
[ -f result/step1/matrix_final.mtx.gz ] || \
	mmutil_filter_row --mtx result/step1/matrix_qc.mtx.gz \
					  --col result/step1/matrix_qc.cols.gz \
					  --row result/step1/matrix.rows.gz \
					  --score NNZ \
					  --cutoff 100 \
					  --col_cutoff 500 \
					  --out result/step1/matrix_final 2> /dev/null
```

