---
title: "Step 2: cell type annotation"
date: "`r Sys.Date()`"
author: Yongjin Park
bibliography: "MS_Ref.bib"
output:
  html_document:
    toc: true
    keep_md: true
    self_contained: true
---


```{r GlobOpt, include = FALSE}
fig.dir = 'Fig/STEP2/'
dir.create(fig.dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_knit$set(eval.after = 'fig.cap')
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = fig.dir)
knitr::opts_chunk$set(fig.width=8, fig.height=8)
options(stringsAsFactors = FALSE)
FIG.CAP = '**Fig.**'
library(data.table)
library(dplyr)
library(patchwork)
library(mmutilR)
library(Matrix)
source("Util-rmd.R")
source("Util-alg.R")
```

```{r echo = FALSE}
.read.marker.file <- function(.file){
    x <- fread(.file, header=F, col.names=c("feat","ct"))
    .xx <- x[, .(feat=list(feat)), by = .(ct)];
    .ret <- .xx$feat; names(.ret) <- .xx$ct;
    .ret
}
```

```{r}
.hdr <- "result/step1/final_matrix"
.data <- fileset.list(.hdr)
```

## 0. Surface Marker Proteins

* nTconv : CD3+, CD4+, CD8-, CD25-/CD127+, CD45RA+/CD45RO-
* mTconv : CD3+, CD4+, CD8-, CD25-/CD127+, CD45RA-/CD45RO+
* nTreg : CD3+, CD4+, CD8-, CD25+/CD127-, CD45RA+/CD45RO-
* mTreg : CD3+, CD4+, CD8-, CD25+/CD127-, CD45RA-/CD45RO+

```{r include = F}
.pos.markers <- .read.marker.file("marker/surface_round1_positive.txt")
.neg.markers <- .read.marker.file("marker/surface_round1_negative.txt")

.mkdir("result/step2")
.marker.hdr <- "result/step2/marker"
.marker.data <- fileset.list(.marker.hdr)

if.needed(.marker.data, {
    .markers <- unique(c(unlist(.pos.markers),
                         unlist(.neg.markers)))
    .marker.data <-
        rcpp_mmutil_copy_selected_rows(.data$mtx,
                                       .data$row,
                                       .data$col,
                                       .markers,
                                       .marker.hdr)
})
```

## 1. Adjust batch-specific displacement using Batch-balancing k-Nearest Neighbour graph

```{r run_bbknn_before_annotation}
.file <- "result/step2/bbknn.rds"
.mkdir(dirname(.file))

if.needed(.file, {

    .cells <-
        fread(.data$col, header=F, col.names = "tag") %>%
        parse.tag()

    .svd <- rcpp_mmutil_svd(.data$mtx, RANK=50, TAKE_LN=T, EM_ITER = 20, NUM_THREADS=16)

    .bbknn <- rcpp_mmutil_bbknn(r_svd_v = .svd$V,
                                r_svd_u = .svd$U,
                                r_svd_d = .svd$D,
                                r_batches = .cells$batch,
                                knn = 50,
                                NUM_THREADS = 16,
                                USE_SINGULAR_VALUES = F)

    saveRDS(.bbknn, .file)
})

.bbknn <- readRDS(.file)
```

```{r annotation_with_bbknn}
.file <- "result/step2/annotation_bbknn.txt.gz"
if.needed(.file, {

    .annot.out <-
        rcpp_mmutil_annotate_columns(
            pos_labels = list(p1=.pos.markers),
            r_neg_labels = list(n1=.neg.markers),
            r_U = .bbknn$U,
            r_D = .bbknn$D,
            r_V = .bbknn$factors.adjusted,
            row_file = .data$row,
            col_file = .data$col,
            EM_TOL = 1e-8,
            EM_ITER = 500,
            TAKE_LN = F)

    .col <- c("tag", "celltype", "prob", "ln.prob")
    names(.annot.out$annotation) <- .col
    annot.dt <- setDT(.annot.out$annotation) %>%
        parse.tag()

    fwrite(annot.dt, .file)
})
```

## 2. Make major cell type annotation based on marker proteins + Leiden clustering

```{r refine_annotation_by_leiden}
.file <- "Tab/step2_cell_type.txt.gz"
if.needed(.file, {
    .cells <-
        fread(.data$col, header=F, col.names = "tag") %>%
        parse.tag()

    .leiden <- run.leiden(.bbknn$knn.adj, .cells$tag, res=1, nrepeat = 100)

    .tab <-
        .leiden %>% 
        left_join(annot.dt) %>%
        na.omit()

    ## identify problematic clusters
    .fraction <- .tab[, .(.N), by = .(celltype, membership, component)]
    .fraction[, Ntot := sum(`N`), by = .(membership, component)]
    .mem.qc <- .fraction[order(`N`, decreasing = T), head(.SD, 1), by = .(membership, component)]

    .remove <- .mem.qc[`N` / `Ntot` < .5, .(membership, component)]

    ## Map cell group membership to cell type
    .ct.map <- .mem.qc[, .(membership, component, celltype)]

    final.cell.type <- .leiden %>%
        anti_join(.remove) %>%
        na.omit() %>% 
        left_join(.ct.map) %>%
        as.data.table() %>%
        parse.tag()
    fwrite(final.cell.type, .file)
})
final.cell.type <- fread(.file)
```

```{r calculate_umap_coordinates}
.file <- "Tab/step2_umap_coord.txt.gz"
if.needed(.file, {
    .umap <- uwot::tumap(.bbknn$factors.adjusted,
                         learning_rate = .1,
                         n_epochs = 2000,
                         n_sgd_threads = 16,
                         init = "laplacian",
                         verbose = T,
                         init_sdev = .01,
                         scale = F)

    .tags <- readLines(.data$col)
    colnames(.umap) <- "UMAP" %&% 1:ncol(.umap)
    umap.dt <- data.table(.umap, tag = .tags)
    fwrite(umap.dt, .file)
})
umap.dt <- fread(.file) %>% 
    left_join(final.cell.type) %>% 
    na.omit()
```

```{r calculate_tsne_coordinates}
.file <- "Tab/step2_tsne_coord.txt.gz"
if.needed(.file, {
    .tsne <- Rtsne::Rtsne(.bbknn$factors.adjusted,
                          num_threads = 16,
                          verbose = T,
                          check_duplicates = F)

    .tags <- readLines(.data$col)
    colnames(.tsne$Y) <- "tSNE" %&% 1:ncol(.tsne$Y)
    tsne.dt <- data.table(.tsne$Y, tag = .tags)
    fwrite(tsne.dt, .file)
})
tsne.dt <- fread(.file) %>%
    left_join(final.cell.type) %>%
    na.omit()
```

### t-UMAP

```{r Fig_umap_celltype, fig.width=9, fig.height=5}
p1 <- .gg.plot(umap.dt[sample(.N)], aes(UMAP1, UMAP2, color=celltype)) +
    ggrastr::rasterise(geom_point(stroke = 0, size=1), dpi=300) +
    theme(legend.position = c(0,0), legend.justification = c(0,0)) +
    ggtitle("cell types") +
    scale_color_brewer("", palette = "Paired")

p2 <- .gg.plot(umap.dt[sample(.N)], aes(UMAP1, UMAP2, color=as.factor(batch))) +
    ggrastr::rasterise(geom_point(stroke = 0, size=1), dpi=300) +
    theme(legend.position = c(0,0), legend.justification = c(0,0)) +
    ggtitle("batch membership") +
    scale_color_brewer("", palette = "Set3")

p3 <- .gg.plot(umap.dt[sample(.N)], aes(UMAP1, UMAP2, color=as.factor(membership))) +
    ggrastr::rasterise(geom_point(stroke = 0, size=1), dpi=300) +
    theme(legend.position = c(0,0), legend.justification = c(0,0)) +
    ggtitle("clustering") +
    scale_color_brewer("", palette = "Set1")

plt <- p1 | p2 | p3
print(plt)
```


```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_umap_celltype.pdf"
.gg.save(filename = .file, plot = plt, width=9, height=5)
```

### tSNE

```{r Fig_tsne_celltype, fig.width=9, fig.height=3}
p1 <-
    .gg.plot(tsne.dt[sample(.N)], aes(tSNE1, tSNE2, color=celltype)) +
    ggrastr::rasterise(geom_point(stroke = 0, size=1), dpi=300) +
    theme(legend.position = c(0,0), legend.justification = c(0,0)) +
    ggtitle("cell types") +
    scale_color_brewer("", palette = "Paired")

p2 <-
    .gg.plot(tsne.dt[sample(.N)], aes(tSNE1, tSNE2, color=as.factor(batch))) +
    ggrastr::rasterise(geom_point(stroke = 0, size=1), dpi=300) +
    theme(legend.position = c(0,0), legend.justification = c(0,0)) +
    ggtitle("batch membership") +
    scale_color_brewer("", palette = "Set3")

p3 <-
    .gg.plot(tsne.dt[sample(.N)], aes(tSNE1, tSNE2, color=as.factor(membership))) +
    ggrastr::rasterise(geom_point(stroke = 0, size=1), dpi=300) +
    theme(legend.position = c(0,0), legend.justification = c(0,0)) +
    ggtitle("clustering") +
    scale_color_brewer("", palette = "Set1")

plt <- p1 | p2 | p3
print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_tsne_celltype.pdf"
.gg.save(filename = .file, plot = plt, width=9, height=3)
```

## 3. Confirm cell type assignment with raw surface marker proteins

```{r constrcut_marker_matrix, echo = FALSE}
marker.raw.mtx <- read.dense(.marker.data$mtx)
.names <- sapply(readLines(.marker.data$row),
                 function(s) unlist(strsplit(s, split="[_]+"))[1],
                 USE.NAMES = F)

rownames(marker.raw.mtx) <- .names
colnames(marker.raw.mtx) <- readLines(.marker.data$col)

.melt <-
    reshape2::melt(marker.raw.mtx) %>%
    filter(!str_detect(`Var1`, "Hashtag")) %>%
    as.data.table()

.sum <- .melt[,
              .(value = sum(value)),
              by = .(Var1, Var2)]

## collapse the same feature names
.dt <-
    dcast(.sum, Var1 ~ Var2, value.var = "value", fill = 0)

marker.raw.mtx <- as.matrix(.dt[, -1])
rownames(marker.raw.mtx) <- unlist(.dt[, 1])
```

Two-dimensional density plot on the raw CD marker concentrations.

```{r Fig_cdmarker_raw, fig.width=6, fig.height=5}
.ct <- c("mTreg","nTreg","mTconv","nTconv")
plt <- plt.scatter.ct.2(.ct, final.cell.type, marker.raw.mtx)
print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_cdmarker_raw.pdf"
.gg.save(filename = .file, plot = plt, width=6, height=5)
```
