---
title: "Step 2: cell type annotation"
date: "`r Sys.Date()`"
author: Yongjin Park
bibliography: "MS_Ref.bib"
output:
  html_document:
    toc: true
    keep_md: true
    self_contained: true
---

```{r GlobOpt, include = FALSE}
################################################################
fig.dir = 'Fig/STEP2/'
dir.create(fig.dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_knit$set(eval.after = 'fig.cap')
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = fig.dir)
knitr::opts_chunk$set(fig.width=8, fig.height=8)
options(stringsAsFactors = FALSE)
FIG.CAP = '**Fig.**'
library(data.table)
library(dplyr)
library(patchwork)
library(mmutilR)
source("Util-rmd.R")
source("Util-alg.R")
.fread <- function(...) fread(..., header=FALSE)
.fwrite <- function(...) fwrite(..., quote=FALSE, sep="\t", row.names = FALSE, col.names = FALSE)
```

```{r include=FALSE}
plt.scatter.ct.2 <- function(.ct.show, .assign.tab, .mtx,
                             m1x = "CD25",
                             m1y = "CD127",
                             m2x = "CD45RO",
                             m2y = "CD45RA") {

    .idx <- data.table(tag = colnames(.mtx)) %>%
        mutate(j = 1:n()) %>%
        left_join(.assign.tab[as.character(celltype) %in% .ct.show]) %>%
        select(j, celltype, prob)

    .dt <- data.table(m1x = .mtx[m1x, .idx$j],
                      m1y = .mtx[m1y, .idx$j],
                      m2x = .mtx[m2x, .idx$j],
                      m2y = .mtx[m2y, .idx$j],
                      prob = .idx$prob,
                      celltype = factor(.idx$celltype, .ct.show))
    .dt <- .dt[celltype %in% .ct.show]
    .med.dt <-
        .dt[m1x > 0 &
            m2x > 0 &
            m1y > 0 &
            m2y > 0,
            .(m1x = median(m1x),
                m1y = median(m1y),
                m2x = median(m2x),
                m2y = median(m2y)),
            by = .(celltype)]

    .add.lab <- function(.plt) {

        .lab <- function(x) num.int(10^x)
        .brk <- seq(0,5)

        .plt +
            theme_classic() +
            theme(axis.text = element_text(size=6)) +
            scale_x_continuous(breaks=.brk, labels=.lab, limits=c(-.1,3.2)) +
            scale_y_continuous(breaks=.brk, labels=.lab, limits=c(-.1,3.2)) +
            theme(axis.text.x = element_text(angle=90, vjust=1, hjust=1)) +
            facet_grid(.~celltype) +
            stat_density_2d(geom = "raster",
                            aes(fill = after_stat(density)),
                            show.legend=F,
                            contour=F) +
            scale_fill_viridis_c() +
            stat_density2d(linewidth=.2, colour="white") +
            geom_point(data=.med.dt, colour=2, pch=3, size=1, stroke=1) +
            geom_abline(slope=1, linewidth=.2, colour="white")
    }
    .aes <- aes(x=log10(m1x), y=log10(m1y))
    p1 <- (ggplot(.dt[m1x > 0 & m1y > 0], .aes) %>% .add.lab) +
        xlab(m1x) + ylab(m1y)
    .aes <- aes(x=log10(m2x), y=log10(m2y))
    p2 <- (ggplot(.dt[m2x > 0 & m2y > 0], .aes) %>% .add.lab) +
        xlab(m2x) + ylab(m2y)
    p1/p2
}
```

# Round 1: Top-level cell type annotation

## Can we distinguish cell types by surface proteins?

```{r echo = FALSE}
.read.marker.file <- function(.file){
    x <- .fread(.file, col.names=c("feat","ct"))
    .xx <- x[, .(feat=list(feat)), by = .(ct)];
    .ret <- .xx$feat; names(.ret) <- .xx$ct;
    .ret
}
```

```{r select_proteins_only, echo = FALSE}
.pos.markers <- .read.marker.file("marker/surface_round1_positive.txt")
.neg.markers <- .read.marker.file("marker/surface_round1_negative.txt")
.markers <- unique(c(unlist(.pos.markers), unlist(.neg.markers)))

.hdr <- "result/step1/matrix_final"
.data <- fileset.list(.hdr)

.mkdir("result/step2")
.prot.hdr <- "result/step2/prot"
.prot.data <- fileset.list(.prot.hdr)

if.needed(.prot.data, {

    .prot.data <-
        rcpp_mmutil_copy_selected_rows(.data$mtx,
                                       .data$row,
                                       .data$col,
                                       .markers,
                                       .prot.hdr)
})
```

* nTconv : CD3+, CD4+, CD8-, CD25-/CD127+, CD45RA+/CD45RO-
* mTconv : CD3+, CD4+, CD8-, CD25-/CD127+, CD45RA-/CD45RO+
* nTreg : CD3+, CD4+, CD8-, CD25+/CD127-, CD45RA+/CD45RO-
* mTreg : CD3+, CD4+, CD8-, CD25+/CD127-, CD45RA-/CD45RO+

Run annotation purely based on marker proteins:

```{r run_annotation_bbknn_adjusted}
.file <- "result/step2/prot_annot.txt.gz"

if.needed(.file, {

    .pos.markers <- .read.marker.file("marker/surface_round1_positive.txt")
    .neg.markers <- .read.marker.file("marker/surface_round1_negative.txt")

    .annot.out <-
        rcpp_mmutil_annotate_columns(
            pos_labels = list(p1=.pos.markers),
            r_neg_labels = list(n1=.neg.markers),
            mtx_file = .prot.data$mtx,
            row_file = .prot.data$row,
            col_file = .prot.data$col,
            EM_TOL = 1e-6,
            EM_ITER = 500,
            TAKE_LN = T)

    annot.dt <- setDT(.annot.out$annotation)
    .col <- c("tag", "celltype", "prob", "ln.prob")
    names(annot.dt) <- .col
    annot.dt[, c("barcode","batch") := tstrsplit(tag, split="_")]
    fwrite(annot.dt, .file)
})

annot.dt <- fread(.file)
```

Two-dimensional density plot on the raw CD marker concentrations.

```{r Fig_round1_cdmarker, fig.width=6, fig.height=5}
prot.raw.mtx <- read.dense(.prot.data$mtx)
.names <- sapply(readLines(.prot.data$row),
                 function(s) unlist(strsplit(s, split="[_]+"))[1],
                 USE.NAMES = F)

rownames(prot.raw.mtx) <- .names
colnames(prot.raw.mtx) <- readLines(.prot.data$col)

.melt <-
    reshape2::melt(prot.raw.mtx) %>%
    filter(!str_detect(`Var1`, "Hashtag")) %>%
    as.data.table()

.sum <- .melt[,
              .(value = sum(value)),
              by = .(Var1, Var2)]

## collapse the same feature names
.dt <-
    dcast(.sum, Var1 ~ Var2, value.var = "value", fill = 0)

prot.raw.mtx <- as.matrix(.dt[, -1])
rownames(prot.raw.mtx) <- unlist(.dt[, 1])

.ct <- c("mTreg","nTreg","mTconv","nTconv")
plt <- plt.scatter.ct.2(.ct, annot.dt, prot.raw.mtx)
print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_round1_cdmarker.pdf"
.gg.save(filename = .file, plot = plt, width=6, height=5)
```

## Will the same classification results hold in transcriptomic data?

```{r select_genes_only, echo = FALSE}
.hdr <- "result/step1/matrix_final"
.data <- fileset.list(.hdr)

.mkdir("result/step2")
.gene.hdr <- "result/step2/gene"
.gene.data <- fileset.list(.gene.hdr)

if.needed(.gene.data, {
    .genes <-
        .fread(.data$row, col.names="feat") %>%
        filter(str_starts(feat, "ENSG")) %>%
        filter(!str_detect(feat, "_MT-")) %>%
        unlist()

    .gene.data <-
        rcpp_mmutil_copy_selected_rows(.data$mtx,
                                       .data$row,
                                       .data$col,
                                       .genes,
                                       .gene.hdr)
})
```

```{r hash_data}
.features <- readLines(.data$row)
.hash <- .features[str_detect(.features, "Hash")]

.hash.hdr <- "result/step1/hash"
.hash.data <- fileset.list(.hash.hdr)

if.needed(.hash.data, {
    .hash.data <-
        rcpp_mmutil_copy_selected_rows(.data$mtx,
                                       .data$row,
                                       .data$col,
                                       .hash,
                                       .hash.hdr)
})

.hash.mtx <- read.dense(.hash.data$mtx)
.hash.argmax <- apply(.hash.mtx, 2, which.max)

.hash.cells <- .hash.data$col %>%
    fread(header=F, col.names="cell") %>%
    (function(x) {
        x[,c("barcode","batch"):=tstrsplit(`cell`,split="_")];
        x[, barcode := gsub(`barcode`, pattern="-[0-9]$", replacement="")];
        x[, hash := .hash.argmax]
        x[, batch := as.integer(`batch`)]
        return(x);
    })

.sample.info <-
    readxl::read_xlsx("data/Hashing list MS Treg project.xlsx", 1) %>%
    rename(Sample = `Cell type`) %>%
    mutate(hash = gsub("#","",`hash`)) %>%
    mutate(hash = as.integer(`hash`)) %>%
    mutate(disease = substr(`subject`, 1, 2))

.hash.info <-
    left_join(.hash.cells, .sample.info)
```


```{r run_bbknn_genes_only}
.file <- "result/step2/gene_bbknn.rds"

if.needed(.file, {

    .batches <- fread(.gene.data$col, col.names="cell", header=F) %>%
        (function(x) {
            x[, c("barcode", "batch") := tstrsplit(cell, split="[_]")];
            x[, barcode := gsub(`barcode`, pattern="-[0-9]$", replacement="")];
            x[, batch := as.integer(`batch`)]
            x;
        }) %>%
        left_join(.hash.info)

    bb <- .batches %>%
        mutate(b = batch %&% "_" %&% disease) %>%
        select(b) %>%
        unlist()

    .bbknn <- rcpp_mmutil_bbknn_svd(.gene.data$mtx, bb,
                                    knn = 30, RANK = 50,
                                    EM_ITER = 20,
                                    TAKE_LN = TRUE,
                                    NUM_THREADS = 16)
    saveRDS(.bbknn, .file)
})
.bbknn <- readRDS(.file)
```

1. Build a batch-balancing kNN graph

2. Leiden graph-based clustering

3. Construct UMAP the kNN backbone graph

```{r bbknn_umap_leiden_by_genes}
.file <- "result/step2/gene_bbknn_leiden.txt.gz"
.cells <- readLines(.gene.data$col)

if.needed(.file, {
    .bbknn.umap <- run.bbknn.umap(.bbknn$knn,
                                  .cells,
                                  symmetrize=F,
                                  min_dist=.1,
                                  spread=5,
                                  res=2)
    fwrite(.bbknn.umap, .file)
})

.bbknn.umap <- fread(.file)

.bbknn.umap <- .bbknn.umap %>%
    left_join(annot.dt, by = "tag") %>%
    filter(celltype != "CD8") %>%
    mutate(membership = as.factor(membership)) %>%
    as.data.table()

## remove cells too far from the centre
## in order to show more cells
.bbknn.umap[, x := scale(umap1)]
.bbknn.umap[, y := scale(umap2)]
.bbknn.umap[, d := sqrt(x^2 + y^2)]

.size <- .bbknn.umap[, .(.N), by = .(membership)]

.bbknn.umap <-
    .size[N > 0, .(membership)] %>%
    left_join(.bbknn.umap, by = "membership") %>%
    filter(`d` < 2.2) %>%
    as.data.table()
```

### Clustering patterns in scRNA-seq generally agrees with the classification results based on surface proteins

```{r Fig_bbknn_gene_only_umap, fig.width=8, fig.height=3}
p1 <-
    .gg.plot(.bbknn.umap, aes(umap1, umap2, color=membership)) +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300)

p2 <-
    .gg.plot(.bbknn.umap, aes(umap1, umap2, color=celltype)) +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300) +
    scale_color_brewer(palette = "Paired")

plt <- p1 | p2
print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_bbknn_gene_only_umap.pdf"
.gg.save(filename = .file, plot = plt, width=8, height=3)
```

However, gene expression patterns may not be sufficient.

```{r Fig_bbknn_gene_only_cluster, fig.width=5, fig.height=2}
.ct.mem <- .bbknn.umap[, .(nc = .N), by = .(celltype, membership)]
.ct.mem[, ntot := sum(`nc`), by = .(membership)]
.ct.mem[, pr := `nc`/`ntot`]

.df <-
    .ct.mem %>%
    mutate(row = celltype, col = membership,
           weight = logit(pmax(pmin(pr, 1-1e-2), 1e-2))) %>%
    order.pair(ret.tab = TRUE)

plt <-
    .gg.plot(.df, aes(col, row, fill=weight, size = `nc`)) +
    xlab("Leiden clustering") +
    ylab("cell type") +
    theme(legend.position = "bottom") +
    geom_point(pch=22) +
    scale_size_continuous("#cells", range=c(0,5)) +
    scale_fill_distiller("proportion",
                         palette="PuRd", direction = 1,
                         labels=function(x) round(sigmoid(x), 2))
print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_bbknn_gene_only_cluster.pdf"
.gg.save(filename = .file, plot = plt, width=5, height=2)
```

## Joint clustering analysis

```{r run_bbknn_full}
.file <- "result/step2/full_bbknn.rds"

if.needed(.file, {

    .batches <- fread(.data$col, col.names="cell", header=F) %>%
        (function(x) {
            x[, c("barcode", "batch") := tstrsplit(cell, split="[_]")];
            x[, barcode := gsub(`barcode`, pattern="-[0-9]$", replacement="")];
            x[, batch := as.integer(`batch`)]
            x;
        }) %>%
        left_join(.hash.info)

    bb <- .batches %>%
        mutate(b = batch %&% "_" %&% disease) %>%
        select(b) %>%
        unlist()

    .bbknn <- rcpp_mmutil_bbknn_svd(.data$mtx, bb,
                                    knn = 30, RANK = 50,
                                    EM_ITER = 20,
                                    TAKE_LN = TRUE,
                                    NUM_THREADS = 16)
    saveRDS(.bbknn, .file)
})
.bbknn <- readRDS(.file)
```

### 1. Run annotation based on the surface marker proteins

```{r run_annotation_bbknn_adjusted_full}
.file <- "result/step2/prot_annot_full.txt.gz"

if.needed(.file, {
    .pos.markers <- .read.marker.file("marker/surface_round1_positive.txt")
    .neg.markers <- .read.marker.file("marker/surface_round1_negative.txt")

    .annot.out <-
        rcpp_mmutil_annotate_columns(
            pos_labels = list(p1=.pos.markers),
            r_neg_labels = list(n1=.neg.markers),
            row_file = .data$row,
            col_file = .data$col,
            r_U = .bbknn$U,
            r_D = .bbknn$D,
            r_V = .bbknn$V,
            EM_TOL = 1e-6,
            EM_ITER = 500)

    annot.dt <- setDT(.annot.out$annotation)
    .col <- c("tag", "celltype", "prob", "ln.prob")
    names(annot.dt) <- .col
    annot.dt[, c("barcode","batch") := tstrsplit(tag, split="_")]
    fwrite(annot.dt, .file)
})
annot.dt <- fread(.file)
```

Two-dimensional density plot on the raw CD marker concentrations.

```{r Fig_round1_cdmarker_full, fig.width=6, fig.height=5}
prot.raw.mtx <- read.dense(.prot.data$mtx)
colnames(prot.raw.mtx) <- readLines(.prot.data$col)

.names <- sapply(readLines(.prot.data$row),
                 function(s) unlist(strsplit(s, split="[_]+"))[1],
                 USE.NAMES = F) %>%
    sapply(gsub, pattern="anti-human", replacement="", USE.NAMES = F) %>%
    sapply(gsub, pattern="^[-]+", replacement="", USE.NAMES = F)

rownames(prot.raw.mtx) <- .names

.ct <- c("mTreg","nTreg","mTconv","nTconv")
plt <- plt.scatter.ct.2(.ct, annot.dt, prot.raw.mtx)
print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_round1_cdmarker_full.pdf"
.gg.save(filename = .file, plot = plt, width=6, height=5)
```

### 2. Followed by cell-cell graph clustering...

```{r run_leiden_full}
.file <- "result/step2/full_bbknn_leiden.txt.gz"
.cells <- readLines(.data$col)

if.needed(.file, {
    .bbknn.umap <- run.bbknn.umap(.bbknn$knn,
                                  .cells,
                                  symmetrize=F,
                                  res=3,
                                  nrepeat=20,
                                  min_dist=.01,
                                  spread=5)
    fwrite(.bbknn.umap, .file)
})

.bbknn.umap <- fread(.file)

.bbknn.umap <- .bbknn.umap %>%
    left_join(annot.dt, by = "tag") %>%
    mutate(membership = as.factor(membership)) %>%
    as.data.table()

## remove cells too far from the centre
## in order to show more cells
.bbknn.umap[, x := scale(umap1)]
.bbknn.umap[, y := scale(umap2)]
.bbknn.umap[, d := sqrt(x^2 + y^2)]

.size <- .bbknn.umap[, .(.N), by = .(membership)]

.bbknn.umap <-
    .size[N > 1300, .(membership)] %>%
    left_join(.bbknn.umap, by = "membership") %>%
    filter(`d` < 2.2) %>%
    as.data.table()
```

```{r Fig_bbknn_full_umap, fig.width=8, fig.height=3}
p1 <-
    .gg.plot(.bbknn.umap[sample(.N)], aes(umap1, umap2, color=membership)) +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300)

p2 <-
    .gg.plot(.bbknn.umap[sample(.N)], aes(umap1, umap2, color=celltype)) +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300) +
    scale_color_brewer(palette = "Paired")

plt <- p1 | p2
print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_bbknn_full_umap.pdf"
.gg.save(filename = .file, plot = plt, width=8, height=3)
```


```{r Fig_bbknn_full_cluster, fig.width=5, fig.height=2}
.ct.mem <- .bbknn.umap[, .(nc = .N), by = .(celltype, membership)]
.ct.mem[, ntot := sum(`nc`), by = .(membership)]
.ct.mem[, pr := `nc`/`ntot`]

.df <-
    .ct.mem %>%
    mutate(row = celltype, col = membership,
           weight = logit(pmax(pmin(pr, 1-1e-2), 1e-2))) %>%
    order.pair(ret.tab = TRUE)

plt <-
    .gg.plot(.df, aes(col, row, fill=weight, size = `nc`)) +
    xlab("Leiden clustering") +
    ylab("cell type") +
    theme(legend.position = "bottom") +
    geom_point(pch=22) +
    scale_size_continuous("#cells", range=c(0,5)) +
    scale_fill_distiller("proportion",
                         palette="PuRd", direction = 1,
                         labels=function(x) round(sigmoid(x), 2))
print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_bbknn_full_cluster.pdf"
.gg.save(filename = .file, plot = plt, width=5, height=2)
```


```{r results="asis", echo=FALSE}
.file <- "result/step2/full_bbknn_leiden.txt.gz"
.dt <- fread(.file) %>%
    left_join(annot.dt, by = "tag") %>%
    as.data.table()
```

* Annotate each cell cluster with the maximally matching cell type

```{r}
## a. Assign cell types to clusters
.tot.size <- .dt[, .(.N), by = .(membership)]
.tot.size[, membership := as.integer(`membership`)]
.size <- .dt[, .(.N), by = .(membership, celltype)]
.size[, membership := as.integer(`membership`)]

.argmax <- .size[order(`N`), tail(.SD, 1), by = .(membership)]
.argmax[, membership := as.integer(`membership`)]
```

* `r length(unique(.argmax$membership))` modules

```{r}
.argmax[, `N.overlap` := `N`]
.argmax[, `N` := NULL]

.size.argmax <- left_join(.argmax, .tot.size) %>% as.data.table()

.file <- "result/step2/full_bbknn_leiden.txt.gz"
celltype.final <-
    fread(.file) %>%
    left_join(.size.argmax) %>%
    as.data.table() %>%
    na.omit()
```

```{r}
## b. Remove small clusters
celltype.final <- celltype.final[`N` > 100]

## c. Remove ambiguous ones
celltype.final <- celltype.final[`N.overlap`/`N` > .5]
```

;* Remove clusters with mixed cell types and retain `r nrow(.size.argmax[N > 100 & N.overlap/N > .5])` clusters

* `r num.int(nrow(celltype.final))` cells.

```{r echo = F}
celltype.final[, c("barcode","batch") := tstrsplit(`tag`, split="_")]
celltype.final[, batch := as.integer(`batch`)]
celltype.final[, barcode := gsub(`barcode`, pattern="-[0-9]+$", replacement="")]
```

Check with Liang's supervised approach.

```{r Fig_overlap_Liang, fig.width=3.2, fig.height=2.8}
.overlap <-
    fread("data/PRDM1/PRDM1_SL.csv.gz") %>%
    left_join(celltype.final) %>%
    na.omit()

.dt <- .overlap[, .(.N), by = .(cell_type_liang, celltype)]
.dt <- .dt[, lab := num.int(`N`)]

.aes <- aes(celltype, cell_type_liang, fill=`N`, label=`lab`)
plt <- .gg.plot(.dt, .aes) +
    geom_tile(linewidth=.1, colour="black") +
    geom_text(size=3) +
    scale_fill_distiller("N", direction=1) +
    theme(axis.text.x = element_text(angle=90, vjust=1, hjust=1)) +
    xlab("CD annotation + Leiden") +
    ylab("Classifier (Liang)")
print(plt)
```

```{r Fig_bbknn_refined_umap, fig.width=12, fig.height=3, results="asis"}
## remove cells too far from the centre
## in order to show more cells
.temp <- fread("data/PRDM1/PRDM1_SL.csv.gz")
.bbknn.umap <- celltype.final %>% left_join(.temp) %>% as.data.table()

.bbknn.umap[, membership := as.factor(`membership`)]
.bbknn.umap[, x := scale(umap1)]
.bbknn.umap[, y := scale(umap2)]
.bbknn.umap[, d := sqrt(x^2 + y^2)]
.bbknn.umap <- .bbknn.umap[`d` < 2.2 & `N` > 1300]

p1 <-
    .gg.plot(.bbknn.umap[sample(.N)], aes(umap1, umap2, color=membership)) +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300) +
    theme(legend.position = "none")

p2 <-
    .gg.plot(.bbknn.umap[sample(.N)], aes(umap1, umap2, color=celltype)) +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300) +
    scale_color_brewer("annotation\n+Leiden", palette = "Paired")

p3 <-
    .gg.plot(na.omit(.bbknn.umap[sample(.N)]), aes(umap1, umap2, color=cell_type_liang)) +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300) +
    scale_color_brewer("supervised", palette = "Paired")

plt <- wrap_plots(p1, p2, p3, nrow=1)
print(plt)

.file <- fig.dir %&% "/Fig_bbknn_refined_umap.pdf"
.gg.save(filename = .file, plot = plt, width=12, height=3)
```


```{r echo = F, results = "asis"}
.file <- "Tab/step2_celltype.txt.gz"
.mkdir("Tab/")
if.needed(.file,
{
    fwrite(celltype.final, .file, sep = "\t", col.names = T)
})
cat("[**DOWNLOAD:** Cell type estimation]("%&% .file %&% ")\n\n")
```

```{r Fig_bbknn_refined_cluster, fig.width=5, fig.height=2, results="asis"}
.ct.mem <- celltype.final[, .(nc = .N), by = .(celltype, membership)]
.ct.mem[, ntot := sum(`nc`), by = .(membership)]
.ct.mem[, pr := `nc`/`ntot`]

.df <-
    .ct.mem %>%
    mutate(row = celltype, col = membership,
           weight = logit(pmax(pmin(pr, 1-1e-2), 1e-2))) %>%
    order.pair(ret.tab = TRUE)

plt <-
    .gg.plot(.df, aes(col, row, size = `nc`)) +
    xlab("Leiden clustering") +
    ylab("cell type") +
    theme(legend.position = "bottom") +
    geom_point(pch=22, fill="black") +
    scale_size_continuous("#cells", range=c(0,5))
print(plt)

.file <- fig.dir %&% "/Fig_bbknn_refined_cluster.pdf"
.gg.save(filename = .file, plot = plt, width=5, height=2)
```


### 3. Confirm by other marker genes/proteins

```{r}
.markers <-
    c("FOXP3", "ID3", "BACH2", "CXCR3", "PRDM1", "SGK1", "TCF7", "LEF1",
      "SELL", "IL2RA", "IL7R", "IKZF2", "CCR6", "CCR4", "CCR7", "CTLA4",
      "HLA-DRA", "CD25", "CD127", "CD183", "CD196",
      "CD197", "CD194", "CD45RA", "CD45RO",
      "HLA") %>%
    unique

.rows <- .fread(.data$row, col.names="r")
.rows[, c("ensembl_gene_id","hgnc_symbol") := tstrsplit(`r`,split="_")]

.markers <- .rows[`hgnc_symbol` %in% .markers]$r

.marker.hdr <- "result/step2/marker/marker"
.mkdir(dirname(.marker.hdr))
.marker.data <- fileset.list(.marker.hdr)
if.needed(.marker.data, {
    .marker.data <-
        rcpp_mmutil_copy_selected_rows(.data$mtx,
                                       .data$row,
                                       .data$col,
                                       .markers,
                                       .marker.hdr)
})
```

```{r include = FALSE}
.tags <- readLines("result/step2/marker/marker.cols.gz")

.ct <- c("mTreg","nTreg","mTconv","nTconv")
.markers <- readLines(.marker.data$row)

.temp <- copy(celltype.final)
.temp[, membership := as.factor(`membership`)]
.temp[, x := scale(umap1)]
.temp[, y := scale(umap2)]
.temp[, d := sqrt(x^2 + y^2)]
.temp <- .temp[`d` < 2.2 & `N` > 1300,
               .(tag, umap1, umap2, celltype)]

marker.dt <-
    t(read.dense(.marker.data$mtx)) %>%
    (function(x) { colnames(x) <- .markers; x }) %>%
    as.data.table() %>%
    mutate(tag = .tags) %>%
    left_join(.temp, by = "tag") %>%
    na.omit() %>%
    as.data.table() %>%
    melt(id.vars = c("tag", "umap1", "umap2","celltype"),
         variable.name = "marker",
         value.name = "expr") %>%
    mutate(celltype = factor(celltype, .ct)) %>%
    as.data.table

plot.marker.umap <- function(.marker) {

    .dt <- marker.dt[marker == .marker]
    ## handle outliers for better visualization
    .cutoff <- quantile(.dt$expr, probs = .95, na.rm=TRUE)
    .cutoff <- max(.cutoff, 2.5)
    .aes <- aes(x = umap1,
                y = umap2,
                colour = expr)
    .dt <- .dt %>%
        mutate(expr = pmin(expr, .cutoff)) %>%
        arrange(expr)
    ret <-
        ggplot(.dt, .aes) +
        ggtitle(.marker) +
        xlab("UMAP 1") + ylab("UMAP 2") +
        theme_minimal() +
        ggrastr::rasterise(geom_point(stroke=.0, size=.7, alpha=.8), dpi=300) +
        theme(legend.key.width = unit(.2, "lines"),
              legend.key.height = unit(.5, "lines"),
              legend.title = element_text(size=6),
              legend.text = element_text(size=5))
    if(str_starts(.marker, "anti_")) {
        ret <- ret + scale_colour_distiller("", direction=1)
    } else {
        ret <- ret + scale_colour_distiller("", palette="RdPu", direction=1)
    }
    return(ret)
}
```

**Note: these plots show expression values without batch adjustment**

```{r Fig_marker, fig.width=3.2, fig.height=2.8, results="asis"}
.markers <- sort(as.character(unique(marker.dt$marker)))
for(g in .markers){
    plt <- plot.marker.umap(g)
    print(plt)
    .file <- fig.dir %&% "/Fig_marker_" %&% g %&% "_umap.pdf"
    .gg.save(filename = .file, plot = plt, width=3.2, height=2.8)
}
```

## Basic statistics for the first round annotation (`r num.int(nrow(celltype.final))` cells)

```{r include=FALSE}
.sum.stat.batch <- function(.dt) {
    .dt %>%
        group_by(batch, disease) %>%
        mutate(P = 100 * N/sum(N)) %>%
        arrange(desc(ct)) %>%
        mutate(Pc = cumsum(P)) %>%
        ungroup() %>%
        group_by(batch) %>%
        mutate(nb = sum(N)) %>%
        ungroup() %>%
        mutate(batch = "#" %&% batch %&% " (N=" %&% num.int(nb) %&% ")")
}
.sum.stat.tot <- function(.dt) {
    .dt %>%
        group_by(disease) %>%
        mutate(P = 100 * N/sum(N)) %>%
        arrange(desc(ct)) %>%
        mutate(Pc = cumsum(P)) %>%
        ungroup()
}
.plt.sum.stat <- function(.stat) {
    .gg.plot(.stat, aes(x=disease, y=P, group=ct, fill=celltype)) +
        facet_grid(.~batch) +
        geom_bar(stat="identity") +
        geom_text(aes(y=Pc, label=num.int(N)), size=2) +
        scale_fill_brewer(palette="Paired") +
        ylab("cell type composition (%)") +
        theme(legend.title=element_blank())
}
```


```{r}
.hash.mtx <- read.dense("result/step1/hash.mtx.gz")
.hash.argmax <- apply(.hash.mtx, 2, which.max)

.cells <- "result/step1/hash.cols.gz" %>%
    fread(header=F, col.names="cell") %>%
    (function(x) {
        x[,c("barcode","batch"):=tstrsplit(`cell`,split="_")];
        x[, barcode := gsub(`barcode`, pattern="-[0-9]$", replacement="")];
        x[, hash := .hash.argmax]
        x[, batch := as.integer(`batch`)]
        return(x);
    })

.sample.info <-
    readxl::read_xlsx("data/Hashing list MS Treg project.xlsx", 1) %>%
    rename(Sample = `Cell type`) %>%
    mutate(hash = gsub("#","",`hash`)) %>%
    mutate(hash = as.integer(`hash`)) %>%
    mutate(disease = substr(`subject`, 1, 2))

.hash.info <-
    left_join(.cells, .sample.info)

.dt <-
    celltype.final %>%
    mutate(batch=as.integer(batch)) %>%
    left_join(.hash.info) %>%
    mutate(ct=factor(celltype, c("nTconv","mTconv","nTreg","mTreg"))) %>%
    na.omit()
```

## Combining all the experiments

```{r Fig_count_stat_tot, fig.height=3, fig.width=6, echo = FALSE, results="asis"}
.stat <-
    .dt[,
        .(N = .N),
        by=.(batch, celltype, ct, disease)] %>%
    .sum.stat.batch
plt <- .plt.sum.stat(.stat) +
    ggtitle("Profiled on all the samples")
print(plt)
.file <- fig.dir %&% "/Fig_count_stat_tot.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=6)
```

```{r Fig_count_merged_stat_tot, fig.height=3, fig.width=2.2, echo = FALSE, results="asis"}
.stat.tot <-
    .dt[,
        .(N = .N),
        by=.(celltype, ct, disease)] %>%
    .sum.stat.tot %>%
    mutate(batch = "(N=" %&% num.int(sum(.stat$N)) %&% ")")
plt <- .plt.sum.stat(.stat.tot) +
    ggtitle("Profiled on all the samples")
print(plt)
.file <- fig.dir %&% "/Fig_count_merged_stat_tot.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=2.2)
```

## Total CD4 samples

```{r Fig_count_stat_cd4, fig.height=3, fig.width=6, echo = FALSE, results="asis"}
.stat <-
    .dt[Sample == "Total CD4T cells",
        .(N = .N),
        by=.(batch, celltype, ct, disease)] %>%
    .sum.stat.batch
plt <- .plt.sum.stat(.stat) +
    ggtitle("Profiled on Total CD4 T-cells")
print(plt)
.file <- fig.dir %&% "/Fig_count_stat_cd4.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=6)
```

```{r Fig_count_merged_stat_cd4, fig.height=3, fig.width=2.2}
.stat <-
    .dt[Sample == "Total CD4T cells", .(N = .N), by=.(celltype, ct, disease)] %>%
    .sum.stat.tot %>%
    mutate(batch = "(N=" %&% num.int(sum(.stat$N)) %&% ")")
plt <- .plt.sum.stat(.stat) +
    ggtitle("Profiled on Total CD4 T-cells")
print(plt)
```

```{r echo = FALSE, results="asis"}
.file <- fig.dir %&% "/Fig_count_merged_stat_cd4.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=2.2)
```

## CD25 enriched samples

```{r Fig_count_stat_cd25enriched, fig.height=3, fig.width=6}
.stat <-
    .dt[Sample == "CD25+ CD4T cells",
        .(N = .N),
        by=.(batch, celltype, ct, disease)] %>%
    .sum.stat.batch
plt <- .plt.sum.stat(.stat) +
    ggtitle("Profiled on the CD25+ enriched")
print(plt)
```

```{r echo = FALSE, results="asis"}
.file <- fig.dir %&% "/Fig_count_stat_cd25enriched.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=6)
```

```{r Fig_count_merged_stat_cd25enriched, fig.height=3, fig.width=2.2}
.stat <-
    .dt[Sample == "CD25+ CD4T cells", .(N = .N), by=.(celltype, ct, disease)] %>%
    .sum.stat.tot %>%
    mutate(batch = "(N=" %&% num.int(sum(.stat$N)) %&% ")")
plt <- .plt.sum.stat(.stat) +
    ggtitle("Profiled on the CD25+ enriched")
print(plt)
```

```{r echo = FALSE, results="asis"}
.file <- fig.dir %&% "/Fig_count_merged_stat_cd25enriched.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=2.2)
```

### PRDM1 short vs. long

```{r include = FALSE}
prdm.thm <-
    theme_minimal() +
    theme(legend.key.width = unit(.2, "lines"),
          legend.key.height = unit(.5, "lines"),
          legend.title = element_text(size=6),
          legend.text = element_text(size=5))
```

```{r incldue = FALSE}
prdm.dt <- fread("data/PRDM1/PRDM1_SL.csv.gz")

.dt <-
    copy(celltype.final) %>%
    left_join(prdm.dt) %>%
    na.omit() %>%
    as.data.table()

.dt[, membership := as.factor(`membership`)]
.dt[, x := scale(umap1)]
.dt[, y := scale(umap2)]
.dt[, d := sqrt(x^2 + y^2)]
.dt <- .dt[`d` < 2.2 & `N` > 1300]
```

```{r Fig_PRDM1_isoform_umap, fig.width=7, fig.height=3}
.aes <- aes(umap1, umap2, colour=pmin(PRDM1_short, 3))

p1 <-
    ggplot(.dt[order(PRDM1_short)], .aes) +
    xlab("UMAP 1") + ylab("UMAP 2") + prdm.thm +
    ggrastr::rasterise(geom_point(stroke=0, size=1), dpi = 300) +
    scale_colour_distiller("PRDM1\nshort",
                           palette="RdPu",
                           direction=1)

.aes <- aes(umap1, umap2, colour = pmin(PRDM1_long, 3))

p2 <-
    ggplot(.dt[order(PRDM1_long)], .aes) +
    xlab("UMAP 1") + ylab("UMAP 2") + prdm.thm +
    ggrastr::rasterise(geom_point(stroke=0, size=1), dpi = 300) +
    scale_colour_distiller("PRDM1\nlong",
                           palette="RdPu",
                           direction=1)

plt <- p1 | p2
print(plt)
```

```{r results="asis", echo=FALSE}
.file <- fig.dir %&% "/Fig_PRDM1_isoform_umap.pdf"
.gg.save(filename = .file, plot = plt, width=7, height=3)
```

### Stratify cells by batches and disease status

```{r include=F}
.dt <- celltype.final %>%
    mutate(batch=as.integer(batch)) %>%
    left_join(.hash.info) %>%
    na.omit()

.dt[, membership := as.factor(`membership`)]
.dt[, x := scale(umap1)]
.dt[, y := scale(umap2)]
.dt[, d := sqrt(x^2 + y^2)]
.dt <- .dt[`d` < 2.2 & `N` > 1300]
```

```{r Fig_bbknn_full_batch, fig.width=8, fig.height=6}
.aes <- aes(umap1, umap2, colour=as.factor(batch))

B <- length(unique(.dt$batch))

plt <-
    .gg.plot(.dt[sample(.N)], .aes) +
    xlab("UMAP 1") + ylab("UMAP 2") +
    ggtitle("stratified by " %&% B %&% " batches") +
    facet_wrap( ~ "#" %&% batch, nrow=2) +
    ggrastr::rasterise(geom_point(stroke=0, size=.7), dpi = 300) +
    scale_color_manual(values=.more.colors(B,.palette="Set1"),
                       guide="none")
print(plt)
```

```{r results="asis", echo=FALSE}
.file <- fig.dir %&% "/Fig_bbknn_full_batch.pdf"
.gg.save(filename = .file, plot = plt, width=8, height=6)
```


```{r Fig_bbknn_full_disease, fig.width=5.5, fig.height=3}
.aes <- aes(umap1, umap2, colour=disease)

B <- length(unique(.dt$disease))

plt <-
    .gg.plot(.dt[sample(.N)], .aes) +
    xlab("UMAP 1") + ylab("UMAP 2") +
    ggtitle("stratified by " %&% B %&% " disease status") +
    facet_wrap( ~ disease, nrow = 1) +
    ggrastr::rasterise(geom_point(stroke=0, size=.7), dpi = 300) +
    scale_color_brewer(guide="none", palette="Set2")
print(plt)
```


```{r results="asis", echo=FALSE}
.file <- fig.dir %&% "/Fig_bbknn_full_disease.pdf"
.gg.save(filename = .file, plot = plt, width=5.5, height=3)
```
