---
title: "Step 4: subtype analysis"
author: Yongjin Park
theme: jekyll-theme-minimal
date: "`r Sys.Date()`"
bibliography: "MS_Ref.bib"
output:
  html_document:
    toc: true
    keep_md: true
    self_contained: true
---


```{r GlobOpt, include = FALSE}
fig.dir <- 'Fig/STEP4/'
dir.create(fig.dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_knit$set(eval.after = 'fig.cap')
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = fig.dir)
knitr::opts_chunk$set(fig.width=8, fig.height=8, results="asis")
options(stringsAsFactors = FALSE)
FIG.CAP <- '**Fig.**'
library(data.table)
library(tidyverse)
library(patchwork)
library(mmutilR)
library(matrixTests)
source("Util-rmd.R")
source("Util-alg.R")
```

```{r helper_functions}
parse.tag <- function(x) {
    x[, c("barcode", "batch") := tstrsplit(tag, split="[_]")];
    x[, barcode := gsub(`barcode`, pattern="-[0-9]$", replacement="")];
    x[, batch := as.integer(`batch`)]
    x;
}
parse.gene <- function(x){
    x[, c("ensembl_gene_id","hgnc_symbol") := tstrsplit(`gene`,split="_")]
    x[, i := 1:.N]
    return(x)
}

bbknn.x <- function(.data, .bbknn, .bbknn.umap, .subset = NULL) {
    .cols <-
        fread(.data$col, header=F, col.names = "tag") %>%
        parse.tag() %>%
        mutate(j = 1:n()) %>%
        left_join(.bbknn.umap) %>%
        na.omit() %>%
        as.data.table()

    .rows <-
        fread(.data$row, col.names="gene", header=F) %>%
        parse.gene()

    if(!is.null(.subset)){
        .rows <- .rows %>%
            filter(hgnc_symbol %in% .subset |
                   ensembl_gene_id %in% .subset)
    }

    vv <- .bbknn$factors.adjusted[.cols$j, , drop = F]
    X <- sweep(.bbknn$U, 2, .bbknn$D, `*`) %*% t(vv)

    ## double scale
    X <- apply(X, 2, scale)
    X.t <- apply(t(X), 2, scale)
    X <- t(X.t)

    ret <- X[.rows$i, , drop = F]
    rownames(ret) <- .rows$hgnc_symbol
    colnames(ret) <- .cols$tag
    return(ret)
}

bbknn.x.melt <- function(.data, .bbknn, .bbknn.umap, .subset = NULL) {

    X <- bbknn.x(.data, .bbknn, .bbknn.umap, .subset)
    ret <- reshape2::melt(X) %>% as.data.table()
    ret[, tag := Var2]
    ret[, gene := Var1]
    ret[, Var1 := NULL]
    ret[, Var2 := NULL]
    ret[, x := value]
    ret[, value := NULL]
    return(ret)
}

take.marker.stats <- function(x, dict){

    require(matrixTests)

    ret <- data.table()
    for(k in unique(dict$membership)){
        tag.k <- .bbknn.umap[membership == k]$tag
        x.k <- x[, colnames(x) %in% tag.k, drop = F]
        y.k <- x[, !(colnames(x) %in% tag.k), drop = F]
        ## wilcox within vs. outside
        wilcox.k <- row_wilcoxon_twosample(x.k, y.k)
        ## t-test
        ttest.k <- row_t_welch(x.k, y.k)
        ret.k <- data.table(
            gene = rownames(x),
            membership = k,
            wilcox.p = wilcox.k$pvalue,
            ttest.t = ttest.k$statistic,
            mean = ttest.k$mean.x,
            mean.outside = ttest.k$mean.y,
            sd = sqrt(ttest.k$var.x),
            sd.outside = sqrt(ttest.k$var.y),
            ttest.p = ttest.k$pvalue)
        ret <- rbind(ret, ret.k)
    }
    return(ret)
}
```

# Results

* Marker genes well-known for mTconv cell subtype classification

```{r}
.markers <-
    c("CCR4","CCR6","CXCR3","CXCR5",
      "ABCA1","GBP4","CDCA7L","ITM2C","NR1D1",
      "CTSH","GATA3","FHIT","CD40LG","C1orf162",
      "MGATA4A","GZMK","IFNGR2",
      "CD183","CD184","CD185","CD196","CD195","CD194",
      "RORC", "TBX21", "HLADR", "CD74", "TCF7", "LEF1",
      "SELL", "CCR7", "CCR8", "IKZF2", "TIGIT", "CD226",
      "BATF", "ANXA2", "BRD9", "HPGD", "LMNA", "TNFRSF4",
      "FOXP3", "FOXP1", "PDCD1", "CD279", "CTLA4", "LAG3",
      "HAVCR2", "CD366", "KLRB1", "FOSL2", "S100A4", "GMAP7",
      "JUN", "IL7R", "MYC", "IL32", "ISG20", "MALAT1",
      "GSDMD", "HDAC1", "GIMAP4", "APOBEC3G", "CD2", "CD28", "CD6",
      "CDKN2A", "CORO1A", "FAS", "FLI1", "GPR25", "MT2A", "KEAP1",
      "IL12RB", "SIRT2", "TNFRSF14", "TRAF3IP3", "IRF2", "PSPH",
      "CD278", "B2M", "RPS26", "MAP1S", "SGK1", "BACH2",
      "HLA-C", "HLA-B", "HLA-E", "HLA-DR", "HLA-DRA", "HLA-DRB1",
      "S1PR4", "KLF2", "SATB1", "TSC22D3", "IL2RA", "CD25") %>%
    unique
```

* Goal: Identify cellular states/subtypes in memory T cells

```{r read_sample_info, inclue = FALSE}
.sample.info <-
    readxl::read_xlsx("data/Hashing list MS Treg project.xlsx", 1) %>%
    mutate(disease=substr(`subject`,1,2)) %>%
    mutate(hash = gsub(pattern="#", replacement="", `hash`)) %>%
    mutate(hash = as.integer(hash)) %>%
    as.data.table()
```

```{r read_hash_info, include = FALSE}
.hash.hdr <- "result/step1/hash"
.hash.data <- fileset.list(.hash.hdr)

.hash.mtx <- read.dense(.hash.data$mtx)
.hash.argmax <- apply(.hash.mtx, 2, which.max)

.hash.cells <- .hash.data$col %>%
    fread(header=F, col.names="tag") %>%
    parse.tag() %>%
    mutate(hash = .hash.argmax)

.hash.info <-
    left_join(.hash.cells, .sample.info)
```

```{r read_cell_annot}
annot.dt <- fread("Tab/step2_celltype.txt.gz")
```

## 1. Memory T conventional

```{r take_subset_mtconv}
.full.data <- fileset.list("result/step1/matrix_final")
.mkdir("result/step4/")
.data <- fileset.list("result/step4/mtconv")

if.needed(.data, {
    .tags <- unique(annot.dt[celltype == "mTconv"]$tag)
    .data <-
        rcpp_mmutil_copy_selected_columns(.full.data$mtx,
                                          .full.data$row,
                                          .full.data$col,
                                          .tags,
                                          "result/step4/mtconv")
})
```

### Clustering cells by batch-balancing k-nearest neighbour graph

```{r run_bbknn_mtconv}
.file <- "result/step4/mtconv_bbknn.rds"
if.needed(.file, {

    .batches <-
        fread(.data$col, col.names="tag", header=F) %>%
        parse.tag() %>%
        left_join(.hash.info)

    bb <- .batches %>%
        mutate(b = batch %&% "_" %&% disease) %>%
        select(b) %>%
        unlist()

    .bbknn <- rcpp_mmutil_bbknn_svd(.data$mtx, bb,
                                    knn = 30, RANK = 50,
                                    EM_ITER = 20,
                                    TAKE_LN = TRUE,
                                    NUM_THREADS = 16)
    saveRDS(.bbknn, .file)
})
.bbknn <- readRDS(.file)
```

```{r bbknn_umap_leiden_mtconv}
.file <- "result/step4/mtconv_bbknn_leiden.txt.gz"
if.needed(.file, {
    .cells <- readLines(.data$col)
    .bbknn.umap <- run.bbknn.umap(.bbknn$knn,
                                  .cells,
                                  symmetrize=T,
                                  min_dist=0.01,
                                  spread=5,
                                  res=.5)
    fwrite(.bbknn.umap, .file)
})

.bbknn.umap <- fread(.file)
.bbknn.umap[, membership := as.factor(membership)]
```

```{r include = F}
.cells <-
    data.table::copy(.bbknn.umap) %>%
    parse.tag() %>%
    left_join(.hash.info) %>%
    select(barcode, batch, hash, subject, tag, membership,
           umap1, umap2, disease)
```

```{r echo = F, results = "asis"}
.file <- "Tab/step4_mtconv_cluster.txt.gz"
.mkdir("Tab/")
if.needed(.file,
{
    fwrite(.cells, .file, sep = "\t", col.names = T)
})
cat("[**DOWNLOAD:** mTconv clustering results]("%&% .file %&% ")\n\n")
```

### What are the cell-cluster-specific marker genes?

```{r find_mtconv_marker_genes}
.mkdir("Tab/")
.file <- "Tab/step4_mtconv_gene_stat.txt.gz"
if.needed(.file, {
    x <- bbknn.x(.data, .bbknn, .bbknn.umap)
    marker.stat <- take.marker.stats(x, .bbknn.umap)
    fwrite(marker.stat, .file, sep = "\t", col.names = T)
})
marker.stat <- fread(.file, sep = "\t")
```

```{r echo = F, results = "asis"}
cat("[**DOWNLOAD:** mTconv marker gene statistics]("%&% .file %&% ")\n\n")
```

### Non-linear embedding to confirm the cell clusters of mTconv cells

```{r Fig_bbknn_mtconv, fig.width=3, fig.height=3}
.lab <-
    .cells[,
           .(umap1=median(umap1), umap2=median(umap2)),
           by = .(membership)]
.cols <- .more.colors(nrow(.lab), nc.pal=12)
plt <-
    .gg.plot(.bbknn.umap, aes(umap1, umap2, color=membership)) +
    xlab("UMAP1") + ylab("UMAP2") +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300) +
    geom_text(aes(label=membership), data=.lab, size=4, color="black") +
    scale_color_manual(values = .cols, guide="none")

print(plt)
.file <- fig.dir %&% "/Fig_bbknn_mtconv.pdf"
.gg.save(filename = .file, plot = plt, width=3, height=3)
```

```{r Fig_bbknn_mtconv_sub, fig.width=3, fig.height=3}
.cols <- .more.colors(10, nc.pal=7, .palette="Set1")
plt <-
    .gg.plot(.cells, aes(umap1, umap2, color=as.factor(subject))) +
    xlab("UMAP1") + ylab("UMAP2") +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300) +
    scale_color_manual(values = .cols, guide="none")

print(plt)
.file <- fig.dir %&% "/Fig_bbknn_mtconv_sub.pdf"
.gg.save(filename = .file, plot = plt, width=3, height=3)
```

```{r select_mtconv_marker_genes}
x.melt <- bbknn.x.melt(.data, .bbknn, .bbknn.umap, .markers)
```

#### Summary heatmap

```{r}
.dt <- x.melt %>% left_join(.cells)
.sum.subj <- .dt[, .(x = median(x)), by = .(gene, subject, membership)]
.sum.subj[, x := scale(x), by = .(gene)]
```

```{r Fig_mtconv_sum_membership, fig.width=1.8, fig.height=6.5}
.sum <-
    .sum.subj[, .(x = median(x)), by = .(gene, membership)] %>%
    mutate(col = `gene`, row = membership, weight = x) %>%
    col.order(1:10, TRUE) %>%
    as.data.table()

plt <-
    .gg.plot(.sum, aes(row, col, fill=pmin(pmax(weight, -1.5), 1.5)))+
    geom_tile(linewidth=.1, color="black") +
    scale_fill_distiller("", palette = "RdBu", direction = -1) +
    theme(legend.key.width = unit(.2,"lines")) +
    theme(legend.key.height = unit(.5,"lines")) +
    xlab("cell clusters") + ylab("features")
print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_mtconv_sum_membership.pdf"
.gg.save(filename = .file, plot = plt, width=1.8, height=6)
```

```{r}
.marker.order <- sort(unique(.sum$col))
```

```{r Fig_mtconv_sum_subj_member, fig.width=5, fig.height=6.5}
.dt <- copy(.sum.subj) %>%
    mutate(gene = factor(`gene`, .marker.order)) %>%
    mutate(t = subject %&% "." %&% membership)

plt <-
    .gg.plot(.dt, aes(`t`, `gene`, fill=pmin(pmax(`x`, -1.5), 1.5))) +
    facet_grid(. ~ membership, space="free", scales="free")+
    geom_tile(linewidth=.1, color="black") +
    scale_fill_distiller("", palette = "RdBu", direction = -1) +
    theme(legend.key.width = unit(.2,"lines")) +
    theme(legend.key.height = unit(.5,"lines")) +
    theme(axis.ticks.x = element_blank()) +
    theme(axis.text.x = element_blank()) +
    xlab("subjects") + ylab("features")

print(plt)
```

```{r results="asis", echo=F}
.file <- fig.dir %&% "/Fig_mtconv_sum_subj_member.pdf"
.gg.save(filename = .file, plot = plt, width=5, height=6)
```

**NOTE** The colors are standardized `log1p` expression across genes and cells.

```{r Fig_mtconv_gene_umap, fig.width=3, fig.height=2.5}
for(g in unique(x.melt$gene)) {
    .dt <- left_join(x.melt[gene == g], .cells)
    .aes <- aes(umap1, umap2, color=pmax(pmin(x, 3), -3))

    plt <-
        .gg.plot(.dt[order(`x`)], .aes) +
        xlab("UMAP1") + ylab("UMAP2") +
        ggrastr::rasterise(geom_point(stroke = 0, size=.7), dpi=300) +
        theme(legend.key.width = unit(.2,"lines")) +
        theme(legend.key.height = unit(.5,"lines")) +
        scale_color_distiller(g, palette = "RdBu", direction = -1) +
        ggtitle(g)

    print(plt)
    .file <- fig.dir %&% "/Fig_mtconv_gene_umap" %&% g %&% ".pdf"
    .gg.save(filename = .file, plot = plt, width=3, height=2.5)
}
```

### Basic statistics

```{r include=FALSE}
.sum.stat.batch <- function(.dt) {
    .dt %>%
        group_by(batch, disease) %>%
        mutate(P = 100 * N/sum(N)) %>%
        arrange(desc(membership)) %>%
        mutate(Pc = cumsum(P)) %>%
        ungroup() %>%
        group_by(batch) %>%
        mutate(nb = sum(N)) %>%
        ungroup() %>%
        mutate(batch="#" %&% batch %&% " (N=" %&% num.int(nb) %&% ")")
}
.sum.stat.tot <- function(.dt) {
    .dt %>%
        group_by(disease) %>%
        mutate(P = 100 * N/sum(N)) %>%
        arrange(desc(membership)) %>%
        mutate(Pc = cumsum(P)) %>%
        ungroup()
}
.plt.sum.stat <- function(.stat, .pal = "Set1") {

    K <- length(unique(.stat$membership))

    .gg.plot(.stat, aes(x=disease, y=P, group=membership, fill=membership)) +
        facet_grid(.~batch) +
        geom_bar(stat="identity") +
        geom_text(aes(y=Pc, label=num.int(N)), size=2) +
        scale_fill_manual(values=.more.colors(K, nc.pal=12)) +
        ylab("clustering membership (%)") +
        theme(legend.title=element_blank())
}
```

```{r Fig_count_mtconv_tot, fig.width=6, fig.height=3}
.stat <-
    .cells[,
           .(N = .N),
           by=.(batch, membership, disease)] %>%
    .sum.stat.batch()

plt <- .plt.sum.stat(.stat) + ggtitle("mTconv")
print(plt)
```

```{r echo = FALSE, results="asis"}
.file <- fig.dir %&% "/Fig_count_mtconv_tot.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=6)
```

```{r Fig_count_merged_mtconv_tot, fig.height=3, fig.width=2.2}
.stat.tot <-
    .cells[,
           .(N = .N),
           by=.(membership, disease)] %>%
    .sum.stat.tot %>%
    mutate(batch = "(N=" %&% num.int(sum(.stat$N)) %&% ")")

plt <- .plt.sum.stat(.stat.tot) + ggtitle("mTconv")
print(plt)
```

```{r echo = FALSE, results="asis"}
.file <- fig.dir %&% "/Fig_count_merged_mtconv_tot.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=2.2)
```


## 2. Memory Treg cells

```{r take_subset_mtreg}
.full.data <- fileset.list("result/step1/matrix_final")
.mkdir("result/step4/")
.data <- fileset.list("result/step4/mtreg")

if.needed(.data, {
    .tags <- unique(annot.dt[celltype == "mTreg"]$tag)
    .data <-
        rcpp_mmutil_copy_selected_columns(.full.data$mtx,
                                          .full.data$row,
                                          .full.data$col,
                                          .tags,
                                          "result/step4/mtreg")
})
```

### Clustering cells by batch-balancing k-nearest neighbour graph

```{r run_bbknn_mtreg}
.file <- "result/step4/mtreg_bbknn.rds"
if.needed(.file, {

    .batches <-
        fread(.data$col, col.names="tag", header=F) %>%
        parse.tag() %>%
        left_join(.hash.info)

    bb <- .batches %>%
        mutate(b = batch %&% "_" %&% disease) %>%
        select(b) %>%
        unlist()

    .bbknn <- rcpp_mmutil_bbknn_svd(.data$mtx, bb,
                                    knn = 30, RANK = 50,
                                    EM_ITER = 20,
                                    TAKE_LN = TRUE,
                                    NUM_THREADS = 16)
    saveRDS(.bbknn, .file)
})
.bbknn <- readRDS(.file)
```


```{r bbknn_umap_leiden_mtreg}
.file <- "result/step4/mtreg_bbknn_leiden.txt.gz"
.cells <- readLines(.data$col)

if.needed(.file, {
    .bbknn.umap <- run.bbknn.umap(.bbknn$knn,
                                  .cells,
                                  symmetrize=T,
                                  min_dist=0.01,
                                  spread=5,
                                  res=.5)
    fwrite(.bbknn.umap, .file)
})

.bbknn.umap <- fread(.file)
.bbknn.umap[, membership := as.factor(membership)]
```

```{r include = F}
.cells <-
    data.table::copy(.bbknn.umap) %>%
    parse.tag() %>%
    left_join(.hash.info) %>%
    select(barcode, batch, hash, subject, tag, membership,
           umap1, umap2, disease)
```

```{r echo = F, results = "asis"}
.file <- "Tab/step4_mtreg_cluster.txt.gz"
.mkdir("Tab/")
if.needed(.file,
{
    fwrite(.cells, .file, sep = "\t", col.names = T)
})
cat("[**DOWNLOAD:** mTreg clustering results]("%&% .file %&% ")\n\n")
```

### What are the cell-cluster-specific marker genes?

```{r find_mtreg_marker_genes}
.mkdir("Tab/")
.file <- "Tab/step4_mtreg_gene_stat.txt.gz"
if.needed(.file, {
    x <- bbknn.x(.data, .bbknn, .bbknn.umap)
    marker.stat <- take.marker.stats(x, .bbknn.umap)
    fwrite(marker.stat, .file, sep = "\t", col.names = T)
})
marker.stat <- fread(.file, sep = "\t")
```

```{r echo = F, results = "asis"}
cat("[**DOWNLOAD:** mTreg marker gene statistics]("%&% .file %&% ")\n\n")
```

### Non-linear embedding to confirm the cell clusters of mTreg cells

```{r Fig_bbknn_mtreg, fig.width=3, fig.height=3}
.lab <-
    .cells[,
           .(umap1=median(umap1), umap2=median(umap2)),
           by = .(membership)]
.cols <- .more.colors(nrow(.lab), nc.pal=12)
plt <-
    .gg.plot(.bbknn.umap, aes(umap1, umap2, color=membership)) +
    xlab("UMAP1") + ylab("UMAP2") +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300) +
    geom_text(aes(label=membership), data=.lab, size=4, color="black") +
    scale_color_manual(values = .cols, guide="none")

print(plt)
.file <- fig.dir %&% "/Fig_bbknn_mtreg.pdf"
.gg.save(filename = .file, plot = plt, width=3, height=3)
```

#### Subject-level variation is not the major source of this clustering pattern.

```{r Fig_bbknn_mtreg_sub, fig.width=3, fig.height=3}
.cols <- .more.colors(10, nc.pal=7, .palette="Set1")
plt <-
    .gg.plot(.cells, aes(umap1, umap2, color=as.factor(subject))) +
    xlab("UMAP1") + ylab("UMAP2") +
    ggrastr::rasterise(geom_point(stroke=0, alpha=.8, size=.7), dpi=300) +
    scale_color_manual(values = .cols, guide="none")
print(plt)
.file <- fig.dir %&% "/Fig_bbknn_mtreg_sub.pdf"
.gg.save(filename = .file, plot = plt, width=3, height=3)
```

#### Confirm cell type identity based on known marker genes?

```{r select_mtreg_marker_genes}
x.melt <- bbknn.x.melt(.data, .bbknn, .bbknn.umap, .markers)
```

```{r}
.dt <- x.melt %>% left_join(.cells)
.sum.subj <- .dt[, .(x = median(x)), by = .(gene, subject, membership)]
.sum.subj[, x := scale(x), by = .(gene)]
```

#### Summary heatmap

```{r Fig_mtreg_sum_membership, fig.width=2, fig.height=6.5}
.sum <-
    .sum.subj[, .(x = median(x)), by = .(gene, membership)] %>%
    mutate(col = `gene`, row = membership, weight = x) %>%
    col.order(1:10, TRUE) %>%
    as.data.table()

plt <-
    .gg.plot(.sum, aes(row, col, fill=pmin(pmax(weight, -1.5), 1.5)))+
    geom_tile(linewidth=.1, color="black") +
    scale_fill_distiller("", palette = "RdBu", direction = -1) +
    theme(legend.key.width = unit(.2,"lines")) +
    theme(legend.key.height = unit(.5,"lines")) +
    xlab("cell clusters") + ylab("features")

print(plt)
.file <- fig.dir %&% "/Fig_mtreg_sum_membership.pdf"
.gg.save(filename = .file, plot = plt, width=2, height=6)
```

```{r}
.marker.order <- sort(unique(.sum$col))
```

```{r Fig_mtreg_sum_subj_member, fig.width=6, fig.height=6.5}
.dt <- copy(.sum.subj) %>%
    mutate(gene = factor(`gene`, .marker.order)) %>%
    mutate(t = subject %&% "." %&% membership)

plt <-
    .gg.plot(.dt, aes(`t`, `gene`, fill=pmin(pmax(`x`, -1.5), 1.5))) +
    facet_grid(. ~ membership, space="free", scales="free")+
    geom_tile(linewidth=.1, color="black") +
    scale_fill_distiller("", palette = "RdBu", direction = -1) +
    theme(legend.key.width = unit(.2,"lines")) +
    theme(legend.key.height = unit(.5,"lines")) +
    theme(axis.ticks.x = element_blank()) +
    theme(axis.text.x = element_blank()) +
    xlab("subjects") + ylab("features")

print(plt)
.file <- fig.dir %&% "/Fig_mtreg_sum_subj_memberhip.pdf"
.gg.save(filename = .file, plot = plt, width=6, height=6)
```

**NOTE** The colors are standardized `log1p` expression across genes and cells.

```{r Fig_mtreg_gene_umap, fig.width=3, fig.height=2.5}
for(g in unique(x.melt$gene)) {
    .dt <- left_join(x.melt[gene == g], .cells)
    .aes <- aes(umap1, umap2, color=pmax(pmin(x, 3), -3))

    plt <-
        .gg.plot(.dt[order(`x`)], .aes) +
        xlab("UMAP1") + ylab("UMAP2") +
        ggrastr::rasterise(geom_point(stroke = 0, size=.7), dpi=300) +
        theme(legend.key.width = unit(.2,"lines")) +
        theme(legend.key.height = unit(.5,"lines")) +
        scale_color_distiller(g, palette = "RdBu", direction = -1) +
        ggtitle(g)

    print(plt)
    .file <- fig.dir %&% "/Fig_mtreg_gene_umap_" %&% g %&% ".pdf"
    .gg.save(filename = .file, plot = plt, width=3, height=2.5)
}
```

```{r Fig_count_mtreg_tot, fig.width=6, fig.height=3}
.stat <-
    .cells[,
           .(N = .N),
           by=.(batch, membership, disease)] %>%
    .sum.stat.batch()

plt <- .plt.sum.stat(.stat) + ggtitle("mTreg")
print(plt)
```

```{r echo = FALSE, results="asis"}
.file <- fig.dir %&% "/Fig_count_mtreg_tot.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=6)
```

```{r Fig_count_merged_mtreg_tot, fig.height=3, fig.width=2.2}
.stat.tot <-
    .cells[,
           .(N = .N),
           by=.(membership, disease)] %>%
    .sum.stat.tot %>%
    mutate(batch = "(N=" %&% num.int(sum(.stat$N)) %&% ")")

plt <- .plt.sum.stat(.stat.tot) + ggtitle("mTreg")
print(plt)
```

```{r echo = FALSE, results="asis"}
.file <- fig.dir %&% "/Fig_count_merged_mtreg_tot.pdf"
.gg.save(filename = .file, plot = plt, height=3, width=2.2)
```


### PRDM1 vs. SGK1 (also standardized across cells)

```{r Fig_mtreg_gene_umap_other, fig.width=3, fig.height=2.5, echo = F}
x.melt <- bbknn.x.melt(.data, .bbknn, .bbknn.umap, c("SGK1", "PRDM1"))

for(g in unique(x.melt$gene)) {
    .dt <- left_join(x.melt[gene == g], .cells)
    .aes <- aes(umap1, umap2, color=pmax(pmin(x, 3), -3))

    plt <-
        .gg.plot(.dt[order(`x`)], .aes) +
        xlab("UMAP1") + ylab("UMAP2") +
        ggrastr::rasterise(geom_point(stroke = 0, size=.7), dpi=300) +
        theme(legend.key.width = unit(.2,"lines")) +
        theme(legend.key.height = unit(.5,"lines")) +
        scale_color_distiller(g, palette = "RdBu", direction = -1) +
        ggtitle(g)

    print(plt)
    .file <- fig.dir %&% "/Fig_mtreg_gene_umap" %&% g %&% ".pdf"
    .gg.save(filename = .file, plot = plt, width=3, height=2.5)
}
```

### What about the PRDM1 long and short isoforms?

```{r Fig_mtreg_PRDM1_iso_umap, fig.width=6, fig.height=2.6, echo = F}
prdm.dt <- fread("data/PRDM1/PRDM1_SL.csv.gz") %>%
    select(-hash)

.dt <-
    copy(.cells) %>%
    left_join(prdm.dt) %>%
    na.omit() %>%
    as.data.table()

.aes <- aes(umap1, umap2, colour=pmin(PRDM1_short, 3))

p1 <-
    .gg.plot(.dt[order(PRDM1_short)], .aes) +
    xlab("UMAP1") + ylab("UMAP2") + ggtitle("shrot") +
    ggrastr::rasterise(geom_point(stroke=0, size=1), dpi = 300) +
    theme(legend.key.width = unit(.2,"lines")) +
    theme(legend.key.height = unit(.5,"lines")) +
    scale_colour_distiller("PRDM1\nshort",
                           palette="RdPu",
                           direction=1)

.aes <- aes(umap1, umap2, colour=pmin(PRDM1_long, 3))

p2 <-
    .gg.plot(.dt[order(PRDM1_long)], .aes) +
    xlab("UMAP1") + ylab("UMAP2") + ggtitle("long") +
    ggrastr::rasterise(geom_point(stroke=0, size=1), dpi = 300) +
    theme(legend.key.width = unit(.2,"lines")) +
    theme(legend.key.height = unit(.5,"lines")) +
    scale_colour_distiller("PRDM1\nlong",
                           palette="RdPu",
                           direction=1)

plt <- p1 | p2
print(plt)
```

```{r results="asis", echo = F}
.file <- fig.dir %&% "/Fig_mtreg_gene_umap_prdm1_iso.pdf"
.gg.save(filename = .file, plot = plt, width=6, height=2.6)
```

### Can we find new marker genes/proteins for mTreg clusters?

```{r select_top_markers}
## Actually, p-values are meaningless because we did clustering based
## on expression and testing the significance of clustering results.
.stat <-
    marker.stat[p.adjust(wilcox.p) < .01 &
                p.adjust(ttest.p) < .01 &
                ttest.t > 0]

.top.markers <-
    .stat[order(`ttest.t`, `sd`, decreasing = T),
          head(.SD, 10),
          by = .(membership)]

top.stat <- .stat[gene %in% .top.markers$gene]
```

