---
title: "Step 3: differential Expression analysis"
date: "`r Sys.Date()`"
author: Yongjin Park
bibliography: "MS_Ref.bib"
---

```{r GlobOpt, include=FALSE}
fig.dir = 'Fig/STEP3/'
knitr::opts_knit$set(eval.after = 'fig.cap')
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = fig.dir)
knitr::opts_chunk$set(fig.width=8, fig.height=8)
options(stringsAsFactors = FALSE)
FIG.CAP = '**Fig.**'

library(tidyverse)
library(data.table)
library(patchwork)
library(mmutilR)
source("Util-rmd.R")
.mkdir(fig.dir)
```

# 1. MS vs. HC at the major cell type level

```{r read_annot}
.hash.hdr <- "result/step1/hash"
.hash.data <- fileset.list(.hash.hdr)
.hash.mtx <- read.dense(.hash.data$mtx)
.hash.argmax <- apply(.hash.mtx, 2, which.max)

.hash.cells <- .hash.data$col %>%
    fread(header=F, col.names="cell") %>%
    (function(x) {
        x[,c("barcode","batch"):=tstrsplit(`cell`,split="_")];
        x[, barcode := gsub(`barcode`, pattern="-[0-9]$", replacement="")];
        x[, hash := .hash.argmax]
        x[, batch := as.integer(`batch`)]
        return(x);
    })

.sample.info <-
    readxl::read_xlsx("data/Hashing list MS Treg project.xlsx", 1) %>%
    rename(Sample = `Cell type`) %>%
    mutate(hash = gsub("#","",`hash`)) %>%
    mutate(hash = as.integer(`hash`)) %>%
    mutate(disease = substr(`subject`, 1, 2))

.hash.info <-
    left_join(.hash.cells, .sample.info)

annot.dt <-
    fread("Tab/step2_celltype.txt.gz") %>%
    left_join(.hash.info) %>%
    na.omit()

sort.col <- function(.mat, .cts, .indvs){

    .fun <- function(ct) paste0(.indvs, "_", ct)

    .samples <-
        lapply(.cts, .fun) %>%
        do.call(what=c) %>%
        sort()

    .idx <- match(.samples, colnames(.mat))
    .idx <- .idx[!is.na(.idx)]
    .mat[, .idx, drop = FALSE]
}

combine.statistics <- function(.result){

    ret <- lapply(names(.result), function(v) {
        .mat <- .result[[v]]
        if(!is.matrix(.mat)) return(NULL)
        as.data.table(reshape2::melt(.mat, value.name = v))
    })

    out <- ret[[1]]
    for(j in 2:length(ret)){
        if(is.null(out)) {
            out <- ret[[j]]
        } else if(!is.null(ret[[j]])) {
            out <- merge(out, ret[[j]])
        }
    }
    return(out)
}

summarize.deg <- function(.dt, tot.cutoff = 3, disease.name = "MS", control.name = "HA") {

    ## adjust potential bias
    .dt[`tot` >= tot.cutoff, b := mean(cfa), by = .(gene, celltype)]
    .dt[, cfa := cfa - b]

    ## .sd <- .dt[`tot` >= tot.cutoff,
    ##            .(min.sd = sd(cfa) / sqrt(.N)),
    ##            by = .(gene, celltype)]
    ## left_join(.sd) %>%

    .dt <- .dt %>%
        filter(`tot` >= tot.cutoff) %>%
        mutate(w = 1/cfa.sd^2) %>%
        as.data.table

    ## average disease effect on the disease subjects
    ADD.dt <-
        .dt[disease == disease.name & `tot` >= tot.cutoff,
               .(ADD = sum(cfa * w) / sum(w),
                 ADD.se = sqrt(1/sum(w))),
               by = .(gene, celltype)]

    ## average disease effect on the control subjects
    ADC.dt <-
        .dt[disease == control.name & `tot` >= tot.cutoff,
               .(ADC = -sum(cfa * w)/ sum(w),
                 ADC.se = sqrt(1/sum(w))),
               by = .(gene, celltype)]

    ## average disease effect
    ADE.dt <-
        .dt[`tot` >= tot.cutoff] %>%
        mutate(cfa.s = if_else(disease == disease.name, 1, -1) * cfa) %>%
        mutate(tot.s = if_else(disease == disease.name, 1, -1) * tot) %>%
        as.data.table %>%
        (function(xx) {
            xx[,
               .(ADE = sum(cfa.s * w) / sum(w),
                 ADE.se = sqrt(1/sum(w)),
                 tot = sum(tot)),
               by = .(gene, celltype)
               ]
        }) %>%
        mutate(z = ADE/ADE.se) %>%
        mutate(pv = 2 * pnorm(abs(z), lower.tail=FALSE)) %>%
        mutate(fwer = p.adjust(pv, "holm")) %>%
        mutate(fdr = p.adjust(pv, "fdr")) %>%
        as.data.table

    left_join(ADE.dt, ADD.dt) %>% left_join(ADC.dt) %>% as.data.table
}
```


```{r estimate_deg_statistics}
.data <- fileset.list("result/step1/matrix_final")
.mkdir("result/step3/deg/")
.deg.data <- fileset.list("result/step3/deg/hc_ms")

if.needed(.deg.data, {
    .deg.data <-
        rcpp_mmutil_copy_selected_columns(.data$mtx,
                                          .data$row,
                                          .data$col,
                                          unique(annot.dt$tag),
                                          "result/step3/deg/hc_ms")
})

.file <- "result/step3/deg/hc_ms.rds"

if.needed(.file, {

    .celltype <-
        annot.dt[, .(tag, celltype)] %>%
        as.data.frame()

    .cell2indv <- annot.dt[, .(tag, subject)] %>%
        unique %>%
        as.data.frame()

    .indv2exp <- .cell2indv %>%
        select(subject) %>%
        mutate(disease = substr(`subject`, 1, 2)) %>%
        as.data.frame()

    .deg.stat <-
        make.cocoa(.deg.data, .celltype, .cell2indv, .indv2exp,
                   .rank = 15, .em.iter = 20, .em.tol = 1e-8, .take.ln = TRUE,
                   knn = 50, impute.by.knn = TRUE, num.threads = 8)

   saveRDS(.deg.stat, .file)
})

.deg.stat <- readRDS(.file)

.cts <- unique(annot.dt$celltype)
.indvs <- unique(annot.dt$subject)

.hc.ms.dt <-
    list(tot = sort.col(.deg.stat$sum, .cts, .indvs),
         cfa = sort.col(.deg.stat$resid.ln.mu, .cts, .indvs),
         cfa.sd = sort.col(.deg.stat$resid.ln.mu.sd, .cts, .indvs)) %>%
    combine.statistics()

.hc.ms.dt[, c("subject","celltype") := tstrsplit(`Var2`,split="_")]
.hc.ms.dt[, disease := substr(`subject`,1,2)]
.hc.ms.dt[, gene := `Var1`]
.hc.ms.dt[, `Var1` := NULL]
.hc.ms.dt[, `Var2` := NULL]

hc.ms.deg <- summarize.deg(.hc.ms.dt)
```

### Comparison with the bulk DEG results

```{r}
plot.bulk.sc <- function(.sc, .bulk, fdr.cutoff = .01, tot.cutoff = 100, qv.cutoff = .2, n.top = 7) {

    .dt <- left_join(.bulk, .sc) %>%
        mutate(sc.z = pmin(pmax(scale(z), -10), 10)) %>%
        na.omit %>%
        arrange(abs(sc.z)) %>%
        as.data.table

    .dt.show <-
        .dt[`fdr` < `fdr.cutoff` & `bulk.qv` < qv.cutoff & `tot` > tot.cutoff] %>%
        group_by(sign(sc.z)) %>%
        top_n(n.top, -log10(pv)) %>%
        ungroup() %>%
        as.data.table

    rr <- cor(.dt$z, .dt$bulk.t, method="spearman")
    r.pv <- cor.test(.dt$z, .dt$bulk.t, method="spearman")$p.value

    .dt.pos <- .dt.show[z > 0]
    .dt.neg <- .dt.show[z < 0]

    .scale.size <-
        scale_size_continuous(range=c(1,2.5), labels = function(x) num.sci(10^(-x)), guide = "none")

    plt <-
        ggplot(.dt, aes(x = bulk.t, y = sc.z)) +
        theme_minimal() +
        xlab("bulk RNA-seq DEG effect") + ylab("scRNA-seq DEG effect") +
        scale_x_continuous(limits = c(-7, 7)) +
        geom_point(aes(size = -log10(pmax(pv, bulk.pv))),
                   stroke=.2, pch=21, colour="gray40") +
        .scale.size +
        geom_vline(xintercept = 0, linewidth = .1) +
        geom_hline(yintercept = 0, linewidth = .1) +
        geom_smooth(aes(bulk.t, sc.z), method="lm", se=FALSE, colour="red", lty=1, formula = y ~ x - 1, linewidth=.5)

    .aes <- aes(label = hgnc_symbol %&% "\n(" %&% num.sci(bulk.pv) %&% ")")

    if(nrow(.dt.pos) > 0) {
        plt <- plt +
            geom_point(data=.dt.pos, pch=21, fill="#ef8a62") +
            ggrepel::geom_label_repel(data=.dt.pos, .aes,
                                      size = 3, vjust=1, hjust=1,
                                      segment.size = .1,
                                      segment.color = "#ef8a62",
                                      box.padding = .1,
                                      label.padding = .1,
                                      nudge_x = -1, nudge_y = -1,
                                      max.overlaps = 10)
    }

    if(nrow(.dt.neg) > 0) {
        plt <- plt +
            geom_point(data=.dt.neg, pch=21, fill="#67a9cf") +
            ggrepel::geom_label_repel(data=.dt.neg, .aes,
                                      size = 3, vjust=0, hjust=0,
                                      segment.size = .1,
                                      segment.color = "#67a9cf",
                                      box.padding = .1,
                                      label.padding = .1,
                                      nudge_x = 1, nudge_y = 1,
                                      max.overlaps = 10)
    }

    plt <- plt +
        ggtitle("Cor=" %&%
                num.round(rr) %&%
                ", p=" %&% num.sci(r.pv) %&%
                ", N=" %&% num.int(nrow(.dt)))

    list(plt=plt, cor = rr, cor.pv = r.pv)
}
```

```{r read_bulk_deg_results}
read.bulk <- function(.file) {
    fread(.file) %>%
        rename(hgnc_symbol = gene_name) %>%
        rename(bulk.t = t, bulk.pv = P.Value, bulk.qv = adj.P.Val) %>%
        dplyr::select(hgnc_symbol, starts_with("bulk"))
}

.file <- "data/DEG/20180513/deg.ms.hc.treg.mem.exvivo.sex_covar.ruv.txt"
bulk.mTreg.dt <- read.bulk(.file)

.file <- "data/DEG/20180513/deg.ms.hc.teff.mem.exvivo.sex_covar.ruv.txt"
bulk.mTconv.dt <- read.bulk(.file)
```

```{r ms_vs_hc_results, results="asis", echo = FALSE}
.mkdir("Tab")
.file <- "Tab/DEG_MS_vs_HC.txt.gz"

if.needed(.file, {
    fwrite(hc.ms.deg, .file, sep="\t")
})
cat("[**DOWNLOAD:** DEG MS vs HC]("%&% .file %&% ")\n\n")
```

#### Memory Treg genes

```{r}
.sc <- hc.ms.deg[celltype == "mTreg" &
                 sign(ADD) == sign(z) &
                 sign(ADC) == sign(z),]

.sc[, c("ensembl_gene_id","hgnc_symbol") := tstrsplit(`gene`, split="_")]

.bulk <- bulk.mTreg.dt

.out <- plot.bulk.sc(.sc,
                     .bulk,
                     fdr.cutoff = .01,
                     qv.cutoff = .2,
                     n.top = 10)
```

#### Correlation `r num.round(.out$cor)` with p-value = `r num.sci(.out$cor.pv)`

```{r Fig_DEG_comparison_mTreg, fig.width=3, fig.height=5, echo = FALSE, results="asis"}
.file <- fig.dir %&% "/Fig_DEG_comparison_mTreg.pdf"
print(.out$plt)
.gg.save(filename = .file, plot = .out$plt, width=3, height=5)
```

#### Memory Tconv genes

```{r}
.sc <- hc.ms.deg[celltype == "mTconv" &
                 sign(ADD) == sign(z) &
                 sign(ADC) == sign(z),]

.sc[, c("ensembl_gene_id","hgnc_symbol") := tstrsplit(`gene`, split="_")]

.bulk <- bulk.mTconv.dt

.out <- plot.bulk.sc(.sc,
                     .bulk,
                     fdr.cutoff = .01,
                     qv.cutoff = .2,
                     n.top = 10)
```

#### Correlation `r num.round(.out$cor)` with p-value = `r num.sci(.out$cor.pv)`

```{r Fig_DEG_comparison_mTconv, fig.width=3, fig.height=5, echo = FALSE, results="asis"}
.file <- fig.dir %&% "/Fig_DEG_comparison_mTconv.pdf"
print(.out$plt)
.gg.save(filename = .file, plot = .out$plt, width=3, height=5)
```

### Found `r num.int(nrow(unique(hc.ms.deg[fwer < .05, .(gene)])))` unique genes strongly perturbed by MS with FWER 5%

* Up-regulated: `r num.int(nrow(unique(hc.ms.deg[fwer < .05 & z > 0, .(gene)])))`

* Down-regulated:  `r num.int(nrow(unique(hc.ms.deg[fwer < .05 & z < 0, .(gene)])))`

* Total pairs of genes and cell types: `r num.int(nrow(hc.ms.deg))`

```{r}
count.deg <- function(.dt, fwer.cutoff = .05) {
    .dt[fwer < fwer.cutoff &
        sign(ADD) == sign(ADE) &
        sign(ADC) == sign(ADE),
        .(n = .N),
        by = .(celltype,
               direction = if_else(z > 0, "up", "down"))
        ] %>%
        mutate(direction = factor(direction, c("up", "down"))) %>%
        group_by(celltype) %>%
        arrange(desc(direction)) %>%
        mutate(nc = cumsum(n)) %>%
        ungroup
}
```

```{r Fig_DEG_count, fig.width=2, fig.height=2, echo = FALSE, results="asis"}
.df <- count.deg(hc.ms.deg)

plt <-
    .gg.plot(.df, aes(x = celltype, y = n, group = direction, fill = direction)) +
    theme(legend.position = c(1,1)) +
    theme(legend.justification = c(1,1)) +
    xlab("cell type") + ylab("# DEG") +
    geom_bar(stat="identity") +
    geom_text(aes(y=nc, label=n)) +
    scale_fill_manual(values = c("#ef8a62","#67a9cf"))

print(plt)
.file <- fig.dir %&% "/Fig_DEG_count.pdf"
.gg.save(filename = .file, plot = plt, width=2, height=2)
```

#### Show genes with FWER $< 10^{-2}$ overlapping with the bulk DEGs

```{r}
show.each.gene <- function(g, .data, .stat, .cutoff = 0) {

    .dt.g <- na.omit(.data[gene == g]) %>% filter(tot > .cutoff)
    .stat.dt <- na.omit(.stat[gene == g]) %>% rename(total = tot)

    .dir <- .stat.dt[which.max(abs(.stat.dt$ADE)), .(ADE)] %>%
        unlist

    if(.dir > 0) {

        .dt <- .dt.g %>%
            left_join(.stat.dt, by = c("celltype", "gene")) %>%
            group_by(disease, celltype) %>%
            arrange(cfa) %>%
            mutate(j = 1:n()) %>%
            ungroup() %>%
            mutate(x = disease %&% "_" %&% j)

    } else {

        .dt <- .dt.g %>%
            left_join(.stat.dt, by = c("celltype", "gene")) %>%
            group_by(disease, celltype) %>%
            arrange(desc(cfa)) %>%
            mutate(j = 1:n()) %>%
            ungroup() %>%
            mutate(x = disease %&% "_" %&% j)

    }

    .dt <- .dt %>%
        mutate(.pv = if_else(is.na(pv), "1", num.sci(pv))) %>%
        mutate(.facet = celltype %&% "\n(" %&% .pv %&% ")") %>%
        mutate(ADD.z = ADD/ADD.se, ADC.z = ADC/ADC.se) %>%
        as.data.table

    .stat.dt <- .dt[, .(celltype,
                        ADE, ADE.se,
                        ADD, ADD.se,
                        ADC, ADC.se,
                        pv, .facet)] %>%
        unique

    .aes <- aes(x = x,
                y = cfa,
                ymin = cfa - 2 * cfa.sd,
                ymax = cfa + 2 * cfa.sd,
                fill = disease)

    .scale.y <- scale_y_continuous(labels = function(x) num.round(exp(x)))

    p1 <-
        .gg.plot(.dt, .aes) +
        theme(legend.position = "top") +
        facet_wrap(~.facet, nrow = 1, scales="free") +
        .scale.y +
        theme(axis.ticks.x = element_blank()) +
        theme(axis.text.x = element_blank()) +
        theme(axis.title.x = element_blank()) +
        geom_hline(yintercept = 0, linewidth = 1, colour = "gray") +
        geom_hline(data = .stat.dt, aes(yintercept = ADE), linewidth = 1, colour = "#F8766D") +
        geom_linerange(linewidth = .2) +
        geom_point(aes(size = tot), pch = 21, stroke = .2) +
        scale_size_continuous("total(k)", range=c(0.1, 2.5), labels = function(x) num.int(x/1e3)) +
        scale_fill_manual(values = c("gray90","#F8766D")) +
        xlab("") + ylab("fold change") + ggtitle(g)

    .temp.1 <- .stat.dt %>%
        dplyr::select(celltype, pv, ADE, ADD, ADC) %>%
        mutate(ADC = - ADC) %>%                  # flip the effect
        melt(id.vars=c("celltype", "pv"), value.name = "mu")

    .temp.2 <- .stat.dt %>%
        dplyr::select(celltype, ADE.se, ADD.se, ADC.se) %>%
        melt(id.vars="celltype", value.name = "se") %>%
        mutate(variable = str_remove(variable, ".se"))

    .temp <- left_join(.temp.1, .temp.2, by = c("celltype", "variable"))

    .aes <- aes(x = variable,
                y = mu,
                ymin = mu - 3 * se,
                ymax = mu + 3 * se)

    p2 <-
        .gg.plot(.temp, .aes) +
        theme(axis.text.x = element_text(angle=90, hjust = 1, vjust= .5)) +
        facet_wrap(~celltype, nrow = 1, scales="free") +
        .scale.y + ylab("fold change") + xlab("") +
        geom_hline(yintercept = 0, lty = 2, linewidth = .5, colour = "black") +
        geom_linerange(linewidth = .2) +
        geom_point(aes(fill = variable, stroke = .2, size = -log10(pv)), pch = 21) +
        scale_fill_brewer(palette = "Greens", guide="none") +
        scale_size_continuous("p-value", range=c(.5, 2.5), breaks=seq(0, 10, 2), labels = function(x) 10^(-x))

    plt <- (p1 / p2) + patchwork::plot_layout(heights = c(4,2))
    return(plt)
}
```

```{r}
.bulk.genes <-
    rbind(bulk.mTreg.dt[bulk.qv < .2, .(hgnc_symbol)],
          bulk.mTconv.dt[bulk.qv < .2, .(hgnc_symbol)]) %>%
    unique() %>%
    unlist()

hc.ms.deg[, c("ensembl_gene_id", "hgnc_symbol") := tstrsplit(`gene`, split="_")]

.genes.show <-
    hc.ms.deg[fwer < 0.05 &
              hgnc_symbol %in% .bulk.genes &
              sign(ADD) == sign(ADE) &
              sign(ADC) == sign(ADE)] %>%
    select(gene) %>%
    unique %>%
    unlist %>%
    as.character
```

```{r Fig_DEG_example, results="asis", echo = FALSE}
for(g in .genes.show){
    .file <- fig.dir %&% "/Fig_DEG_example_" %&% g %&% ".pdf"
    if.needed(.file, {
        plt <- show.each.gene(g, .hc.ms.dt, hc.ms.deg)
        .gg.save(filename = .file, cat.link = FALSE, plot = plt, width=5, height=3)
    })
    cat("[" %&% g %&% "](" %&% .file %&% ") ")
}
cat("\n\n\n")
```

### Additional genes/proteins

```{r}
.add <- c("CD279","PDCD1","LAG3","CD223","HAVCR2","TIM3","CD366","TOX","TOX2")

additional.genes <-
    hc.ms.deg[hgnc_symbol %in% .add, .(gene)] %>%
    unique() %>%
    unlist(use.names=F)
```

```{r Fig_DEG_example2, results="asis", echo = FALSE}
for(g in additional.genes){
    .file <- fig.dir %&% "/Fig_DEG_example_" %&% g %&% ".pdf"
    if.needed(.file, {
        plt <- show.each.gene(g, .hc.ms.dt, hc.ms.deg)
        .gg.save(filename = .file, cat.link = FALSE, plot = plt, width=5, height=3)
    })
    cat("[" %&% g %&% "](" %&% .file %&% ") ")
}
cat("\n\n\n")
```

### Appendix: PRDM1 short vs. long

```{r}

.total.cols <- fread(.deg.data$col, col.names = "tag", header=F)

## 1. Write down PRDM1 short and long as .mtx
.hdr <- "result/step3/deg/prdm1_hc_ms"
if.needed(fileset.list(.hdr), {

    .dt <- fread("data/PRDM1/PRDM1_SL.csv.gz")
    .dt[, tag := barcode %&% "-1" %&% "_" %&% `batch`]

    .out <- .dt[(PRDM1_short + PRDM1_long > 0) &
                `tag` %in% .total.cols$tag,
                .(tag, PRDM1_short, PRDM1_long)]


    .mat <- .out[, .(PRDM1_short, PRDM1_long)] %>%
        as.matrix() %>%
        t() %>% 
        Matrix::Matrix(sparse=TRUE)

    mmutilR::write.sparse(.mat,
                          c("PRDM1_short", "PRDM1_long"),
                          .out$tag,
                          .hdr)
})


prdm1.data <- fileset.list(.hdr)

.file <- "result/step3/deg/prdm1_hc_ms.rds"

if.needed(.file, {

    .pca <- rcpp_mmutil_svd(mtx_file = .deg.data$mtx,
                            RANK = 15,
                            TAKE_LN = T)

    .prdm1.cols <- fread(prdm1.data$col, col.names = "tag", header=F)

    V <- .pca$V[match(.prdm1.cols$tag, .total.cols$tag),
              , drop = F]


    .prdm1.cols[, c("barcode", "batch") := tstrsplit(`tag`, split="_")]
    .prdm1.cols[, barcode := gsub(pattern="-1", replacement = "", `barcode`)]
    .prdm1.cols[, batch := as.integer(batch)]

    .matched <- .prdm1.cols %>%
        left_join(annot.dt) %>% 
        left_join(.hash.info)

    .stat <- rcpp_mmutil_aggregate(mtx_file = prdm1.data$mtx, 
                                   row_file = prdm1.data$row,
                                   col_file = prdm1.data$col, 
                                   r_cols = .matched$tag,
                                   r_indv = .matched$subject,
                                   r_annot = .matched$celltype,
                                   r_lab_name = c("mTreg","mTconv","nTreg","nTconv"),
                                   r_trt = .matched$disease,
                                   r_V = V, knn = 50, IMPUTE_BY_KNN = T,
                                   NUM_THREADS = 16)

    saveRDS(.stat, .file)
})

.deg.stat <- readRDS(.file)

.cts <- unique(annot.dt$celltype)
.indvs <- unique(annot.dt$subject)

.hc.ms.dt <-
    list(tot = sort.col(.deg.stat$sum, .cts, .indvs),
         cfa = sort.col(.deg.stat$resid.ln.mu, .cts, .indvs),
         cfa.sd = sort.col(.deg.stat$resid.ln.mu.sd, .cts, .indvs)) %>%
    combine.statistics()

.hc.ms.dt[, c("subject","celltype") := tstrsplit(`Var2`,split="_")]
.hc.ms.dt[, disease := substr(`subject`,1,2)]
.hc.ms.dt[, gene := `Var1`]
.hc.ms.dt[, `Var1` := NULL]
.hc.ms.dt[, `Var2` := NULL]

hc.ms.deg <- summarize.deg(.hc.ms.dt)

```


```{r Fig_SL_DEG_PRDM1_long, results="asis", echo=F, fig.width=5, fig.height=3}
.file <- fig.dir %&% "/Fig_SL_DEG_PRDM1_long.pdf"
plt <- show.each.gene("PRDM1_long", .hc.ms.dt, hc.ms.deg)
print(plt)
.gg.save(filename = .file, cat.link = T, plot=plt, width=5, height=3)
```


```{r Fig_SL_DEG_PRDM1_short, results="asis", echo=F, fig.width=5, fig.height=3}
.file <- fig.dir %&% "/Fig_SL_DEG_PRDM1_short.pdf"
plt <- show.each.gene("PRDM1_short", .hc.ms.dt, hc.ms.deg)
print(plt)
.gg.save(filename = .file, cat.link = T, plot=plt, width=5, height=3)
```

