---
title: "Step 1: merge CITE-seq data followed and exploratory data analysis"
author: Yongjin Park
theme: jekyll-theme-minimal
date: "`r Sys.Date()`"
bibliography: "reference.bib"
---


```{r GlobOpt, include = FALSE}
################################################################
fig.dir = 'Fig/STEP1/'
dir.create(fig.dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_knit$set(eval.after = 'fig.cap')
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = fig.dir)
knitr::opts_chunk$set(fig.width=8, fig.height=8)
options(stringsAsFactors = FALSE)
FIG.CAP = '**Fig.**'
library(data.table)
library(dplyr)
library(patchwork)
library(mmutilR)
source("Util-rmd.R")
```

# 1. Merge all the available batches into one file set

## Combine both protein and RNA features

```{r merge_data}
.mkdir("result/step1")
.out.hdr <- "result/step1/matrix"
.data <- fileset.list(.out.hdr)

if.needed(.data, {
    bats <- hdrs <- ("data/batch_" %&% 3:7)
    .data <-
        rcpp_mmutil_merge_file_sets(r_headers=hdrs,
                                    r_batches=3:7,
                                    r_mtx=str_c(hdrs, ".mtx.gz"),
                                    r_row=str_c(hdrs, ".feature.tsv.gz"),
                                    r_col=str_c(hdrs, ".barcode.tsv.gz"),
                                    output=.out.hdr)
    if.needed(.data$mtx %&% ".index", rcpp_mmutil_build_index(.data$mtx))
    .mkdir(.out.hdr)
})
```

```{r}
.info <- rcpp_mmutil_info(.data$mtx)
```

* Total `r num.int(.info$max.col)` cells

* Total `r num.int(.info$max.row)` features

* Total  `r num.int(.info$max.elem)` non-zero elements


## Take RNA-seq data with basic Q/C steps

```{r take_rna_data}
.genes <-
    fread(.data$row, col.names="feat", header=F) %>%
    filter(!str_starts(feat, "anti_")) %>% 
    filter(!str_starts(feat, "MT-")) %>%
    unlist()
.mt.genes <- 
    fread(.data$row, col.names="feat", header=F) %>%
    filter(str_starts(feat, "MT-")) %>%
    unlist()
.rna.hdr <- "result/step1/rna"
.mt.hdr <- "result/step1/mt"
.rna.data <- fileset.list(.rna.hdr)
.mt.data <- fileset.list(.mt.hdr)

if.needed(.rna.data, {
    .mkdir(.rna.hdr)
    .rna.data <-
        rcpp_mmutil_copy_selected_rows(.data$mtx,
                                       .data$row,
                                       .data$col,
                                       .genes,
                                       .rna.hdr)
})

if.needed(.mt.data, {
    .mkdir(.mt.hdr)
    .mt.data <-
        rcpp_mmutil_copy_selected_rows(.data$mtx,
                                       .data$row,
                                       .data$col,
                                       .mt.genes,
                                       .mt.hdr)
})
```


```{r compute_scores}
.dt.rna <-
    rcpp_mmutil_compute_scores(.rna.data$mtx, .rna.data$row, .rna.data$col) %>%
    (function(x) setDT(x$col))
.dt.mt <- 
    rcpp_mmutil_compute_scores(.mt.data$mtx, .mt.data$row, .mt.data$col) %>%
    (function(x) setDT(x$col))
```

```{r run_bbknn}
.file <- "result/step1/bbknn/rna.rds"
.mkdir(dirname(.file))
if.needed(.file, {
    .batches <- fread(.rna.data$col, col.names="cell", header=F) %>%
        (function(x) {
            x[, c("barcode", "batch") := tstrsplit(cell, split="[_]")];
            x$batch
        })
    .bbknn <- rcpp_mmutil_bbknn_pca(.rna.data$mtx,
                                    .batches,
                                    knn=100, RANK=3,
                                    TAKE_LN = TRUE)
    saveRDS(.bbknn, .file)
})

.bbknn <- readRDS(.file)
.factors <- .bbknn$factors.adjusted
```

# 2. Cell Q/C by the number of non-zero elements and mictochondrial activities

```{r cell_score_cutoff}
.file <- "result/step1/qc_table.txt.gz"

if.needed(.file, {

    .qc.dt <-
        merge(.dt.rna, .dt.mt, by="name", suffix=c(".nonmt", ".mt")) %>% 
        mutate(mito.frac = 100 * sum.mt / (sum.nonmt + sum.mt)) %>%
        as.data.table

    .temp <- .factors %>%
        as.data.table %>%
        (function(x) { colnames(x) <- "PC" %&% 1:ncol(x); x })

        .qc.dt <- cbind(.qc.dt, .temp) %>% as.data.table %>% na.omit
        set.seed(1)
        .x <- .qc.dt$mito.frac
        .kmeans <- kmeans(.x, centers=2, nstart=100)
        k.valid <- which.max(.kmeans$size)
        mito.cutoff <- max(.x[.kmeans$cluster == k.valid])
        .y <- .qc.dt$nnz.nonmt
        .kmeans <- kmeans(log10(.y), centers=2, nstart=100)
        k.valid <- which.max(.kmeans$size)
        nnz.cutoff <- max(.y[.kmeans$cluster != k.valid])

        .dat <- .qc.dt[, .(PC1, PC2, PC3)] %>%
            as.matrix %>%
            apply(MARGIN = 2, scale)

        .kmeans <- kmeans(.dat, centers=2, nstart=100)
        k.valid <- which(apply(.kmeans$centers, 1, function(x) all(abs(x) < 2)))
        pc.valid <- which(.kmeans$cluster %in% k.valid)

        .qc.dt <- .qc.dt %>%
            mutate(j = 1:n()) %>%
            mutate(qc = if_else(nnz.nonmt > nnz.cutoff &
                                mito.frac < mito.cutoff &
                                j %in% pc.valid,
                                "pass",
                                "fail")) %>%
            dplyr::select(-j) %>%
            as.data.table

        fwrite(.qc.dt, .file, col.names = TRUE, row.names = FALSE)
})

.qc.dt <- fread(.file, header = TRUE)
```

```{r qc_colours}
.qc.cols <- c("#8856a7", "gray70")
```

```{r fig_cell_nnz, results="asis", fig.width=6, fig.height=3}
plt <-
    .gg.plot(.qc.dt, aes(x=nnz.nonmt)) +
    geom_histogram(aes(fill=qc), bins=100, colour="black", size=.1) +
    scale_x_continuous("number of genes detected in each cell") +
    scale_fill_manual(values = .qc.cols) +
    ylab("number of cells") +
    theme(legend.position="none")
print(plt)
.file <- fig.dir %&% "/Fig_cell_nnz.pdf"
.gg.save(filename = .file, plot = plt, width=6, height=3)
```

```{r fig_cell_mito, results="asis", fig.width=6, fig.height=3}
.lab <- function(x) num.round(pmax(0,exp(x)-1))
plt <-
    .gg.plot(.qc.dt, aes(x=log(1 + mito.frac))) +
    geom_histogram(aes(fill=qc), bins=100, colour="black", size=.1) +
    scale_fill_manual(values = .qc.cols) +
    scale_x_continuous("% mitochondrial activity", labels = .lab) +
    ylab("count of cells") +
    theme(legend.position="none")
print(plt)
.file <- fig.dir %&% "/Fig_cell_mito.pdf"
.gg.save(filename = .file, plot = plt, width=6, height=3)
```

* Keep `r num.int(nrow(.qc.dt[qc == "pass"]))` cells, discard `r num.int(nrow(.qc.dt[qc == "fail"]))` cells

```{r Fig_svd_batch, results="asis", fig.width=2, fig.height=2, echo = FALSE}
.temp <- .qc.dt[qc == "pass"]
.temp[, c("barcode","batch") := tstrsplit(`name`, split="_")]
plt <-
    .gg.plot(.temp[sample(nrow(.temp))], aes(x=PC1, y=PC2)) +
    geom_point(aes(fill=batch), size = 1, pch = 21, stroke=0, show.legend=FALSE) +
    scale_fill_brewer(palette = "Spectral")
print(plt)
.file <- fig.dir %&% "/Fig_svd_batch.pdf"
.gg.save(filename = .file, plot = plt, width=2, height=2)
```

## Matrix data after Q/C: `r num.int(nrow(.qc.dt[qc == "pass"]))` cells

```{r select_qc_cells}
.qc.hdr <- "result/step1/matrix_qc"
.qc.data <- fileset.list(.qc.hdr)
.qc.cells <- .qc.dt[qc == "pass", .(name)] %>% unlist
if.needed(.qc.data,{
    .qc.data <-
        rcpp_mmutil_copy_selected_columns(.data$mtx,
                                          .data$row,
                                          .data$col,
                                          .qc.cells,
                                          .qc.hdr)
})
```

```{r}
.info <- rcpp_mmutil_info(.qc.data$mtx)
```

* Total `r num.int(.info$max.col)` cells

* Total `r num.int(.info$max.row)` features

* Total  `r num.int(.info$max.elem)` non-zero elements

## Feature Q/C to remove features with too many zeros

```{r feature_qc_data}
.feature.scores <- 
    rcpp_mmutil_compute_scores(.qc.data$mtx, .qc.data$row, .qc.data$col) %>%
    (function(x) setDT(x$row))
.valid.features <- .feature.scores[nnz > 100, .(name)] %>% unlist
.final.hdr <- "result/step1/matrix_final"
.final.data <- fileset.list(.final.hdr)
if.needed(.final.data, {
    .final.data <-
        rcpp_mmutil_copy_selected_rows(.qc.data$mtx,
                                       .qc.data$row,
                                       .qc.data$col,
                                       .valid.features,
                                       .final.hdr)
})
```

```{r}
.info <- rcpp_mmutil_info(.final.data$mtx)
```

* Total `r num.int(.info$max.col)` cells

* Total `r num.int(.info$max.row)` features

* Total  `r num.int(.info$max.elem)` non-zero elements

